---
title: Licenses and Inspections
subtitle: Exploratory Data Analysis
date: today
author:
  - name: Angel Rutherford
  - name: Ixchel Ramirez
  - name: Tess Vu
    email:
      - tessavu@proton.me
      - tessavu@upenn.edu
affiliation:
  - name: University of Pennsylvania
    department: Urban Spatial Analytics (MUSA)
    city: Philadelphia
    state: PA
    url: https://www.design.upenn.edu/urban-spatial-analytics
format:
  html:
    code-fold: show
    toc: true
    toc_float: true
    toc-expand: true
    smooth-scroll: true
    embed-resources: true
    title-block-style: default
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Potential Project?

**Research Question:** What is the probability that a building will be classified as High-Risk/Deficient within the next 12 months?

**Method:** The question I posed has a binary outcome so a logistic classification regression is ideal.

**Data Scope:** From 2002 onward, but there are gaps until 2010, and sparseness between 2010 and 2020.

# Setup and Initial Data Loading

```{r setup}
# Packages
# Spatial
library(sf)

# Data wrangling & visualization (tidyverse loads dplyr, tidyr, stringr,
# lubridate, ggplot2, readr, etc.)
library(tidyverse)

# Tables
library(knitr)

# File paths (data untracked; download from OpenDataPhilly
building_certs_path <- file.path("../data", "license_inspection", "BUILDING_CERTS.csv")
dor_parcel_path <- file.path("../data", "DOR_Parcel", "DOR_Parcel.shp")

# Chunk options
#opts_chunk$set(fig.path = "..results/", dev = "png")
```

```{r load-and-shape}
certs_data <- read_csv(building_certs_path, show_col_types = FALSE) %>%
  mutate(
    inspectiondate = ymd_hms(inspectiondate),
    expirationdate = ymd_hms(expirationdate),
    inspection_year = year(inspectiondate)
  )

cat("\nBUILDING_CERTS Dimension:", dim(certs_data))
cat("\n\nBUILDING_CERTS Columns:\n\n")
print(lapply(certs_data, class))
```

```{r table-view}
certs_data
```

# Certification Types

```{r cert-types}
cert_type_counts <- certs_data %>%
  count(buildingcerttype, sort = TRUE)

cert_type_plot <- cert_type_counts %>%
  slice_head(n = 10) %>%
  ggplot(aes(x = reorder(buildingcerttype, n), y = n)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(
    title = "Top Ten Building Certification Types",
    x     = "Certification Type",
    y     = "Count of Certifications"
  ) +
  theme_minimal()

cert_type_plot
```

# Inspection Results

```{r inspection-results}
# Collapse similar statuses into Safe / Failure buckets.
certs_data <- certs_data %>%
  mutate(
    result_clean = case_match(
      inspectionresult,
      "Certified"        ~ "Safe",
      "Safe with Repair" ~ "Safe",
      "Unsafe"           ~ "Failure",
      "Deficient"        ~ "Failure",
      .default = inspectionresult
    )
  )

result_counts <- certs_data %>%
  count(result_clean)

result_plot <- ggplot(result_counts, aes(x = result_clean, y = n, fill = result_clean)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Distribution of Final Inspection Results",
    x     = "Inspection Result",
    y     = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

result_plot
```

# Temporal Coverage

```{r temporal-coverage}
year_plot <- ggplot(certs_data, aes(x = inspection_year)) +
  geom_histogram(binwidth = 1, fill = "darkblue", color = "white") +
  labs(
    title = "Number of Certifications by Inspection Year",
    x     = "Inspection Year",
    y     = "Count of Certifications"
  ) +
  theme_minimal()

year_plot
```

# Potential Target Variable

```{r target-variable}
certs_data <- certs_data %>%
  mutate(
    days_until_expiration = as.numeric(
      difftime(expirationdate, inspectiondate, units = "days")
    )
  )

duration_plot <- ggplot(certs_data, aes(x = days_until_expiration)) +
  geom_histogram(bins = 30, fill = "darkgreen", color = "white") +
  labs(
    title = "Distribution of Certification Duration in Days",
    x     = "Certification Duration (Days)",
    y     = "Count"
  ) +
  theme_minimal()

duration_plot
```

# Spatial Analysis

```{r spatial-join}
opa_parcels <- st_read(dor_parcel_path) %>%
  drop_na(c(addr_sourc, addr_std)) %>%
  select(addr_sourc, addr_std, parcel)

certs_data <- certs_data %>%
  drop_na(address) %>%
  mutate(addr_std = address)

merged_by_addr_std <- inner_join(
  opa_parcels, certs_data,
  by = "addr_std", relationship = "many-to-many"
)
```

```{r spatial-viz-fail-rate}
district_risk <- merged_by_addr_std %>%
  group_by(council_district) %>%
  summarise(
    total_certs   = n(),
    failure_count = sum(result_clean == "Failure"),
    failure_rate  = failure_count / total_certs,
    .groups = "drop"
  ) %>%
  arrange(desc(failure_rate))

district_risk_plot <- ggplot(
  district_risk,
  aes(x = reorder(council_district, failure_rate), y = failure_rate)
) +
  geom_bar(stat = "identity", fill = "firebrick") +
  labs(
    title = "Failure Rate by City Council District",
    x     = "Council District",
    y     = "Failure Rate (Proportion of Total Certs)"
  ) +
  theme_minimal()

district_risk_plot
```

```{r contractor-performance}
top_contractors <- merged_by_addr_std %>%
  count(contractorname, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(contractorname)

contractor_data <- merged_by_addr_std %>%
  filter(
    contractorname %in% top_contractors,
    !is.na(days_until_expiration)
  )

contractor_duration_plot <- ggplot(
  contractor_data,
  aes(
    x = reorder(contractorname, days_until_expiration, FUN = median),
    y = days_until_expiration
  )
) +
  geom_boxplot(fill = "gold") +
  coord_flip() +
  labs(
    title = "Certification Duration by Top 10 Contractors",
    x     = "Contractor Name",
    y     = "Certification Duration (Days)"
  ) +
  theme_minimal()

contractor_duration_plot
```
