# Spatial Analysis: Tract-Rail Station Interactions

## Introduction

This section performs comprehensive spatial analysis to understand bus-rail interactions at the census tract level. We calculate distances from tract centroids to regional rail stations, identify tracts within rail catchment areas, and test whether rail proximity affects bus ridership patterns.

**Key Questions:**
1. Do tracts near rail stations have lower bus ridership (substitution effect)?
2. Or do they have higher bus ridership (feeder effect)?
3. Does rail proximity affect post-pandemic bus ridership recovery?
4. Which tracts are underserved by rail and depend heavily on bus service?

## Load Spatial Data

```{r load-tract-geometry}
# Load Pennsylvania census tract geometry for SEPTA service area.
tract_geo <- st_read("data/pa_tracts/pa_tracts.shp", quiet = TRUE)

# Filter to five-county SEPTA region using FIPS codes.
septa_county_fips <- c("017", "029", "045", "091", "101")

tract_septa <- tract_geo %>%
  filter(COUNTYFP %in% septa_county_fips) %>%
  # Transform to consistent CRS for distance calculations (meters).
  st_transform(32618)

cat("SEPTA Region Census Tracts:\n")
cat(sprintf("Total Tracts: %s\n", n_distinct(tract_septa$GEOID)))
cat(sprintf("Counties: %s\n", paste(unique(tract_septa$COUNTYFP), collapse = ", ")))

# Check tract geometry.
cat("\nTract Geometry Summary:\n")
print(st_geometry_type(tract_septa) %>% table())
```

## Calculate Tract Centroids

```{r calculate-centroids}
# Calculate geographic centroids for each census tract.
tract_centroids <- tract_septa %>%
  st_centroid()

cat("Tract Centroids Calculated:\n")
cat(sprintf("Number of Centroids: %s\n", nrow(tract_centroids)))
cat(sprintf("CRS: %s\n", st_crs(tract_centroids)$input))

# Verify centroid calculation.
cat("\nFirst 5 Centroids:\n")
tract_centroids %>%
  select(GEOID, COUNTYFP, geometry) %>%
  head(5) %>%
  st_drop_geometry() %>%
  cbind(
    Lon = st_coordinates(head(tract_centroids, 5))[, 1],
    Lat = st_coordinates(head(tract_centroids, 5))[, 2]
  ) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Prepare Regional Rail Station Points

```{r prepare-rail-stations}
# Aggregate regional rail station data to unique station locations.
rr_stations_unique <- df_rr_clean %>%
  group_by(Station, Line, Latitude, Longitude) %>%
  summarize(
    Total_Weekday_Boards = sum(Boards[Service_Ty == "Weekday"], na.rm = TRUE),
    Total_Weekend_Boards = sum(Boards[Service_Ty == "Saturday" | Service_Ty == "Sunday"], na.rm = TRUE),
    Years_Active = n_distinct(Year),
    .groups = "drop"
  ) %>%
  filter(Total_Weekday_Boards > 0)

cat("Unique Regional Rail Stations:\n")
cat(sprintf("Total Stations: %s\n", nrow(rr_stations_unique)))
cat(sprintf("Total Lines: %s\n", n_distinct(rr_stations_unique$Line)))

# Convert rail stations to sf object with same CRS as tracts.
rr_stations_sf <- rr_stations_unique %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>%
  st_transform(32618)

cat("\nRegional Rail Stations Spatial Object Created:\n")
print(st_geometry_type(rr_stations_sf) %>% table())
```

## Calculate Distance to Nearest Rail Station

```{r distance-to-rail}
# Calculate distance from each tract centroid to nearest rail station.
dist_matrix <- st_distance(tract_centroids, rr_stations_sf)

# Find minimum distance for each tract.
tract_rail_distance <- tract_centroids %>%
  mutate(
    # Distance in meters.
    Dist_to_Rail_m = apply(dist_matrix, 1, min),
    # Convert to miles.
    Dist_to_Rail_mi = Dist_to_Rail_m / 1609.34,
    # Convert to kilometers.
    Dist_to_Rail_km = Dist_to_Rail_m / 1000,
    # Find which station is nearest.
    Nearest_Station_Index = apply(dist_matrix, 1, which.min)
  ) %>%
  # Join nearest station information.
  mutate(
    Nearest_Station = rr_stations_unique$Station[Nearest_Station_Index],
    Nearest_Line = rr_stations_unique$Line[Nearest_Station_Index]
  ) %>%
  # Drop geometry for easier manipulation.
  st_drop_geometry()

cat("Distance to Rail Calculations Complete:\n")
cat(sprintf("Mean Distance: %.2f miles\n", mean(tract_rail_distance$Dist_to_Rail_mi)))
cat(sprintf("Median Distance: %.2f miles\n", median(tract_rail_distance$Dist_to_Rail_mi)))
cat(sprintf("Min Distance: %.2f miles\n", min(tract_rail_distance$Dist_to_Rail_mi)))
cat(sprintf("Max Distance: %.2f miles\n", max(tract_rail_distance$Dist_to_Rail_mi)))

# Display summary statistics.
cat("\nDistance Distribution Quantiles (miles):\n")
quantile(tract_rail_distance$Dist_to_Rail_mi, probs = seq(0, 1, 0.1)) %>%
  as.data.frame() %>%
  rename("Distance (miles)" = ".") %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Classify Tracts by Rail Accessibility

```{r classify-rail-access}
# Create categorical rail accessibility classification.
tract_rail_distance <- tract_rail_distance %>%
  mutate(
    Rail_Access = case_when(
      Dist_to_Rail_mi <= 0.5 ~ "Very Close (≤0.5 mi)",
      Dist_to_Rail_mi <= 1.0 ~ "Close (0.5-1 mi)",
      Dist_to_Rail_mi <= 2.0 ~ "Moderate (1-2 mi)",
      Dist_to_Rail_mi <= 5.0 ~ "Far (2-5 mi)",
      TRUE ~ "Very Far (>5 mi)"
    ),
    Rail_Access = factor(
      Rail_Access, 
      levels = c("Very Close (≤0.5 mi)", "Close (0.5-1 mi)", 
                 "Moderate (1-2 mi)", "Far (2-5 mi)", "Very Far (>5 mi)")
    ),
    # Binary indicator for rail proximity.
    Near_Rail = ifelse(Dist_to_Rail_mi <= 1.0, 1, 0)
  )

cat("Rail Accessibility Classification:\n")
tract_rail_distance %>%
  count(Rail_Access) %>%
  mutate(Pct = n / sum(n) * 100) %>%
  kable(digits = 1, col.names = c("Rail Access Category", "N Tracts", "Percent")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Join Distance Metrics to Ridership Data

```{r join-distance-ridership}
# Join distance metrics to main ridership dataframe.
df_clean_with_rail <- df_clean %>%
  left_join(
    tract_rail_distance %>% 
      select(GEOID, Dist_to_Rail_mi, Dist_to_Rail_km, 
             Rail_Access, Near_Rail, Nearest_Station, Nearest_Line),
    by = c("Census_Tract_ID" = "GEOID")
  )

cat("Ridership Data Enhanced with Rail Distance:\n")
cat(sprintf("Total Observations: %s\n", comma(nrow(df_clean_with_rail))))
cat(sprintf("Observations with Rail Distance: %s\n", 
            comma(sum(!is.na(df_clean_with_rail$Dist_to_Rail_mi)))))
cat(sprintf("Missing Rail Distance: %s\n", 
            comma(sum(is.na(df_clean_with_rail$Dist_to_Rail_mi)))))

# Check first few rows.
cat("\nFirst 10 Rows with Rail Distance:\n")
df_clean_with_rail %>%
  filter(!is.na(Dist_to_Rail_mi)) %>%
  select(Census_Tract_ID, Year, On_, Dist_to_Rail_mi, Rail_Access, Nearest_Station) %>%
  head(10) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

---

# Visualization: Spatial Patterns

## Map: Distance to Nearest Rail Station

```{r map-distance-to-rail}
#| fig-dpi: 300
#| fig-height: 10
#| fig-width: 12

# Create choropleth map showing distance to rail.
tract_with_distance <- tract_septa %>%
  left_join(
    tract_rail_distance %>% select(GEOID, Dist_to_Rail_mi, Rail_Access),
    by = "GEOID"
  )

# Transform back to WGS84 for plotting.
tract_with_distance_wgs84 <- tract_with_distance %>%
  st_transform(4326)

rr_stations_wgs84 <- rr_stations_sf %>%
  st_transform(4326)

# Create map.
distance_map <- ggplot() +
  geom_sf(data = tract_with_distance_wgs84, 
          aes(fill = Dist_to_Rail_mi), 
          color = "white", linewidth = 0.1) +
  geom_sf(data = rr_stations_wgs84, 
          color = "#E74C3C", size = 2, shape = 17) +
  scale_fill_viridis_c(
    option = "magma", 
    direction = -1,
    name = "Distance\n(miles)",
    labels = comma
  ) +
  labs(
    title = "Distance to Nearest Regional Rail Station by Census Tract",
    subtitle = "Red triangles = Regional Rail stations | Darker tracts = farther from rail",
    caption = "Data: SEPTA Regional Rail Stations & Census Tracts"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

distance_map
```

## Map: Rail Accessibility Categories

```{r map-rail-access}
#| fig-dpi: 300
#| fig-height: 10
#| fig-width: 12

# Create categorical map showing rail accessibility zones.
access_map <- ggplot() +
  geom_sf(data = tract_with_distance_wgs84, 
          aes(fill = Rail_Access), 
          color = "white", linewidth = 0.1) +
  geom_sf(data = rr_stations_wgs84, 
          color = "black", size = 2.5, shape = 17) +
  scale_fill_manual(
    values = c(
      "Very Close (≤0.5 mi)" = "#27AE60",
      "Close (0.5-1 mi)" = "#F39C12",
      "Moderate (1-2 mi)" = "#E67E22",
      "Far (2-5 mi)" = "#E74C3C",
      "Very Far (>5 mi)" = "#8E44AD"
    ),
    name = "Rail\nAccessibility"
  ) +
  labs(
    title = "Rail Accessibility Classification by Census Tract",
    subtitle = "Black triangles = Regional Rail stations | Color indicates distance band",
    caption = "Data: SEPTA Regional Rail Stations & Census Tracts"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

access_map
```

---

# Analysis: Rail Distance Effects on Bus Ridership

## Ridership by Rail Accessibility

```{r ridership-by-access}
#| fig-dpi: 300
#| fig-height: 7
#| fig-width: 12

# Calculate mean ridership by rail accessibility category.
ridership_by_access <- df_clean_with_rail %>%
  filter(Day_of_Week == "Weekday", !is.na(Rail_Access)) %>%
  group_by(Rail_Access, Year) %>%
  summarize(
    Mean_On = mean(On_, na.rm = TRUE),
    Median_On = median(On_, na.rm = TRUE),
    N_Tracts = n_distinct(Census_Tract_ID),
    .groups = "drop"
  )

# Create line plot showing trends by rail accessibility.
access_trend_plot <- ggplot(ridership_by_access, 
                            aes(x = Year, y = Mean_On, color = Rail_Access, group = Rail_Access)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c(
      "Very Close (≤0.5 mi)" = "#27AE60",
      "Close (0.5-1 mi)" = "#F39C12",
      "Moderate (1-2 mi)" = "#E67E22",
      "Far (2-5 mi)" = "#E74C3C",
      "Very Far (>5 mi)" = "#8E44AD"
    ),
    name = "Rail Accessibility"
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Mean Weekday Bus Ridership by Rail Accessibility (2014-2023)",
    subtitle = "Tracts closer to rail show different ridership patterns and recovery trajectories",
    x = "Year",
    y = "Mean Weekday Boardings per Tract",
    caption = "Data: SEPTA Bus Ridership & Regional Rail Stations"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

access_trend_plot
```

## Distribution: Ridership vs Distance

```{r ridership-distance-scatter}
#| fig-dpi: 300
#| fig-height: 10
#| fig-width: 10

# Create scatter plot for 2023 ridership versus distance.
ridership_2023 <- df_clean_with_rail %>%
  filter(Day_of_Week == "Weekday", Year == 2023, !is.na(Dist_to_Rail_mi))

# Calculate correlation.
dist_ridership_cor <- cor(ridership_2023$Dist_to_Rail_mi, 
                          ridership_2023$On_, 
                          use = "complete.obs")

# Create hex plot.
distance_scatter <- ggplot(ridership_2023, 
                          aes(x = Dist_to_Rail_mi, y = On_ + 1)) +
  geom_hex(bins = 50) +
  geom_smooth(method = "loess", se = TRUE, color = "#E74C3C", linewidth = 1.3) +
  scale_fill_viridis_c(option = "plasma", trans = "log10", labels = comma, name = "Tract\nCount") +
  scale_y_log10(labels = comma) +
  annotate("text", x = max(ridership_2023$Dist_to_Rail_mi) * 0.7, y = 10,
           label = sprintf("Correlation: %.3f", dist_ridership_cor),
           size = 5, fontface = "bold") +
  labs(
    title = "Bus Ridership vs Distance to Nearest Rail Station (2023)",
    subtitle = "Each hex represents multiple census tracts | LOESS curve shows relationship",
    x = "Distance to Nearest Rail Station (miles)",
    y = "Weekday Boardings (Log Scale)",
    caption = "Data: SEPTA 2023 Weekday Bus Ridership"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14))

distance_scatter

cat(sprintf("\nCorrelation: Distance to Rail vs Ridership = %.4f\n", dist_ridership_cor))
cat("Interpretation: Negative = substitution | Positive = complementarity\n")
```

## Box Plot: Ridership Distribution by Distance Band

```{r ridership-boxplot-distance}
#| fig-dpi: 300
#| fig-height: 7
#| fig-width: 12

# Create box plot showing ridership distribution by distance band.
distance_boxplot <- ridership_2023 %>%
  filter(On_ > 0) %>%
  ggplot(aes(x = Rail_Access, y = On_ + 1, fill = Rail_Access)) +
  geom_boxplot(outlier.alpha = 0.3, outlier.size = 1) +
  scale_y_log10(labels = comma) +
  scale_fill_manual(
    values = c(
      "Very Close (≤0.5 mi)" = "#27AE60",
      "Close (0.5-1 mi)" = "#F39C12",
      "Moderate (1-2 mi)" = "#E67E22",
      "Far (2-5 mi)" = "#E74C3C",
      "Very Far (>5 mi)" = "#8E44AD"
    )
  ) +
  labs(
    title = "Bus Ridership Distribution by Rail Accessibility (2023)",
    subtitle = "Box plots show median, quartiles, and outliers | Log scale for clarity",
    x = "Rail Accessibility Category",
    y = "Weekday Boardings (Log Scale)",
    caption = "Data: SEPTA 2023 Weekday Bus Ridership | Positive ridership only"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

distance_boxplot
```

## Statistical Test: Rail Proximity Effect

```{r statistical-test}
# Perform t-test comparing near versus far tracts.
near_rail_ridership <- ridership_2023 %>%
  filter(Near_Rail == 1) %>%
  pull(On_)

far_rail_ridership <- ridership_2023 %>%
  filter(Near_Rail == 0) %>%
  pull(On_)

t_test_result <- t.test(near_rail_ridership, far_rail_ridership)

cat("T-Test: Near Rail (≤1 mile) vs Far Rail (>1 mile)\n")
cat("==================================================\n\n")
cat(sprintf("Near Rail Mean: %.2f boardings\n", mean(near_rail_ridership)))
cat(sprintf("Far Rail Mean: %.2f boardings\n", mean(far_rail_ridership)))
cat(sprintf("Difference: %.2f boardings\n", mean(near_rail_ridership) - mean(far_rail_ridership)))
cat(sprintf("\nt-statistic: %.4f\n", t_test_result$statistic))
cat(sprintf("p-value: %.4f\n", t_test_result$p.value))
cat(sprintf("95%% CI: [%.2f, %.2f]\n", t_test_result$conf.int[1], t_test_result$conf.int[2]))

if (t_test_result$p.value < 0.05) {
  cat("\nResult: STATISTICALLY SIGNIFICANT difference\n")
  if (mean(near_rail_ridership) > mean(far_rail_ridership)) {
    cat("Interpretation: Tracts near rail have HIGHER bus ridership (feeder effect)\n")
  } else {
    cat("Interpretation: Tracts near rail have LOWER bus ridership (substitution effect)\n")
  }
} else {
  cat("\nResult: No statistically significant difference\n")
  cat("Interpretation: Rail proximity does not significantly affect bus ridership\n")
}
```

---

# Recovery Analysis: Rail Proximity Effects

## Recovery by Distance Band

```{r recovery-by-distance}
# Calculate recovery ratios by rail accessibility.
recovery_by_access <- df_clean_with_rail %>%
  filter(Day_of_Week == "Weekday", Year %in% c(2019, 2023), !is.na(Rail_Access)) %>%
  group_by(Census_Tract_ID, Rail_Access) %>%
  summarize(
    Ridership_2019 = sum(On_[Year == 2019]),
    Ridership_2023 = sum(On_[Year == 2023]),
    .groups = "drop"
  ) %>%
  filter(Ridership_2019 > 0) %>%
  mutate(Recovery_Ratio = Ridership_2023 / Ridership_2019)

cat("Recovery Analysis by Rail Accessibility:\n")
recovery_by_access %>%
  group_by(Rail_Access) %>%
  summarize(
    N_Tracts = n(),
    Mean_Recovery = mean(Recovery_Ratio),
    Median_Recovery = median(Recovery_Ratio),
    .groups = "drop"
  ) %>%
  kable(digits = 3, 
        col.names = c("Rail Access", "N Tracts", "Mean Recovery", "Median Recovery")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Recovery Distribution by Rail Proximity

```{r recovery-distribution-rail}
#| fig-dpi: 300
#| fig-height: 7
#| fig-width: 12

# Create violin plot showing recovery distribution.
recovery_violin <- ggplot(recovery_by_access, 
                          aes(x = Rail_Access, y = Recovery_Ratio, fill = Rail_Access)) +
  geom_violin(alpha = 0.7, draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_hline(yintercept = 1.0, linetype = "dashed", color = "black", linewidth = 0.8) +
  scale_fill_manual(
    values = c(
      "Very Close (≤0.5 mi)" = "#27AE60",
      "Close (0.5-1 mi)" = "#F39C12",
      "Moderate (1-2 mi)" = "#E67E22",
      "Far (2-5 mi)" = "#E74C3C",
      "Very Far (>5 mi)" = "#8E44AD"
    )
  ) +
  labs(
    title = "Recovery Ratio Distribution by Rail Accessibility (2023 vs 2019)",
    subtitle = "Horizontal lines show quartiles | Dashed line = full recovery",
    x = "Rail Accessibility Category",
    y = "Recovery Ratio (2023 / 2019)",
    caption = "Data: SEPTA Weekday Bus Ridership"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

recovery_violin
```

---

# Buffer Analysis: Rail Station Catchment Areas

## Create Rail Station Buffers

```{r create-buffers}
# Create buffer zones around rail stations.
rail_buffers_half_mile <- rr_stations_sf %>%
  st_buffer(dist = 804.67)

rail_buffers_one_mile <- rr_stations_sf %>%
  st_buffer(dist = 1609.34)

cat("Rail Station Buffer Zones Created:\n")
cat(sprintf("Half-Mile Buffers: %s\n", nrow(rail_buffers_half_mile)))
cat(sprintf("One-Mile Buffers: %s\n", nrow(rail_buffers_one_mile)))
```

## Spatial Join: Tracts Within Buffers

```{r spatial-join-buffers}
# Identify tracts within half-mile buffer.
tracts_within_half_mile <- tract_septa %>%
  st_join(
    rail_buffers_half_mile %>% select(Station, Line, Total_Weekday_Boards),
    join = st_intersects
  ) %>%
  filter(!is.na(Station)) %>%
  group_by(GEOID) %>%
  slice_max(Total_Weekday_Boards, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  st_drop_geometry() %>%
  select(GEOID, Station_Half_Mile = Station, Line_Half_Mile = Line)

# Identify tracts within one-mile buffer.
tracts_within_one_mile <- tract_septa %>%
  st_join(
    rail_buffers_one_mile %>% select(Station, Line, Total_Weekday_Boards),
    join = st_intersects
  ) %>%
  filter(!is.na(Station)) %>%
  group_by(GEOID) %>%
  slice_max(Total_Weekday_Boards, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  st_drop_geometry() %>%
  select(GEOID, Station_One_Mile = Station, Line_One_Mile = Line)

cat("Tracts Within Rail Station Buffers:\n")
cat(sprintf("Tracts within 0.5 miles of rail: %s (%.1f%%)\n", 
            n_distinct(tracts_within_half_mile$GEOID),
            n_distinct(tracts_within_half_mile$GEOID) / n_distinct(tract_septa$GEOID) * 100))
cat(sprintf("Tracts within 1.0 miles of rail: %s (%.1f%%)\n", 
            n_distinct(tracts_within_one_mile$GEOID),
            n_distinct(tracts_within_one_mile$GEOID) / n_distinct(tract_septa$GEOID) * 100))
```

## Enhanced Ridership Data with Buffer Indicators

```{r enhanced-ridership-buffers}
# Join buffer indicators to ridership data.
df_clean_with_rail_buffers <- df_clean_with_rail %>%
  left_join(tracts_within_half_mile, by = c("Census_Tract_ID" = "GEOID")) %>%
  left_join(tracts_within_one_mile, by = c("Census_Tract_ID" = "GEOID")) %>%
  mutate(
    Within_Half_Mile = ifelse(!is.na(Station_Half_Mile), 1, 0),
    Within_One_Mile = ifelse(!is.na(Station_One_Mile), 1, 0)
  )

cat("Enhanced Dataset with Buffer Indicators:\n")
cat(sprintf("Total Observations: %s\n", comma(nrow(df_clean_with_rail_buffers))))
cat(sprintf("Observations within 0.5 miles: %s (%.1f%%)\n", 
            comma(sum(df_clean_with_rail_buffers$Within_Half_Mile)),
            mean(df_clean_with_rail_buffers$Within_Half_Mile) * 100))
cat(sprintf("Observations within 1.0 miles: %s (%.1f%%)\n", 
            comma(sum(df_clean_with_rail_buffers$Within_One_Mile)),
            mean(df_clean_with_rail_buffers$Within_One_Mile) * 100))
```

## Map: Rail Station Buffers and Tracts

```{r map-buffers}
#| fig-dpi: 300
#| fig-height: 10
#| fig-width: 12

# Transform buffers to WGS84 for plotting.
rail_buffers_half_mile_wgs84 <- rail_buffers_half_mile %>% st_transform(4326)
rail_buffers_one_mile_wgs84 <- rail_buffers_one_mile %>% st_transform(4326)

# Create buffer map.
buffer_map <- ggplot() +
  geom_sf(data = tract_with_distance_wgs84, 
          fill = "gray95", color = "gray70", linewidth = 0.1) +
  geom_sf(data = rail_buffers_one_mile_wgs84, 
          fill = "#3498DB", alpha = 0.2, color = NA) +
  geom_sf(data = rail_buffers_half_mile_wgs84, 
          fill = "#E74C3C", alpha = 0.3, color = NA) +
  geom_sf(data = rr_stations_wgs84, 
          color = "black", size = 2.5, shape = 17) +
  labs(
    title = "Regional Rail Station Catchment Areas",
    subtitle = "Red = 0.5 mile buffer | Blue = 1.0 mile buffer | Black triangles = stations",
    caption = "Data: SEPTA Regional Rail Stations & Census Tracts"
  ) +
  theme_void() +
  theme(plot.title = element_text(face = "bold", size = 14))

buffer_map
```

---

# Regression Analysis: Distance to Rail Effect

## Simple Linear Model

```{r simple-lm}
# Filter to 2023 weekday data for cross-sectional analysis.
model_data_2023 <- df_clean_with_rail_buffers %>%
  filter(Day_of_Week == "Weekday", Year == 2023, !is.na(Dist_to_Rail_mi))

# Estimate simple linear regression.
lm_simple <- lm(log(On_ + 1) ~ Dist_to_Rail_mi, data = model_data_2023)

cat("Simple Linear Model: log(Ridership) ~ Distance to Rail\n")
cat("======================================================\n\n")
summary(lm_simple)

cat("\nInterpretation:\n")
cat(sprintf("Coefficient: %.4f\n", coef(lm_simple)[2]))
if (coef(lm_simple)[2] < 0) {
  cat("Negative coefficient indicates tracts farther from rail have LOWER ridership\n")
  cat("This suggests bus-rail COMPLEMENTARITY (feeder effect)\n")
} else {
  cat("Positive coefficient indicates tracts farther from rail have HIGHER ridership\n")
  cat("This suggests bus-rail SUBSTITUTION (competition effect)\n")
}
```

## Multiple Regression with County Fixed Effects

```{r multiple-regression}
# Estimate model with county fixed effects.
lm_county_fe <- lm(
  log(On_ + 1) ~ Dist_to_Rail_mi + County_Name, 
  data = model_data_2023
)

cat("\nMultiple Regression: log(Ridership) ~ Distance + County FE\n")
cat("==========================================================\n\n")
summary(lm_county_fe)

cat("\nKey Findings:\n")
cat(sprintf("Distance Coefficient: %.4f\n", coef(lm_county_fe)[2]))
cat(sprintf("R-squared: %.4f\n", summary(lm_county_fe)$r.squared))
cat(sprintf("Adjusted R-squared: %.4f\n", summary(lm_county_fe)$adj.r.squared))
```

## Binary Proximity Model

```{r binary-proximity-model}
# Estimate model using binary proximity indicator.
lm_binary <- lm(
  log(On_ + 1) ~ Within_One_Mile + County_Name, 
  data = model_data_2023
)

cat("\nBinary Proximity Model: log(Ridership) ~ Within 1 Mile + County FE\n")
cat("==================================================================\n\n")
summary(lm_binary)

cat("\nInterpretation:\n")
if (coef(lm_binary)[2] > 0) {
  cat(sprintf("Tracts within 1 mile of rail have %.1f%% HIGHER ridership\n", 
              (exp(coef(lm_binary)[2]) - 1) * 100))
} else {
  cat(sprintf("Tracts within 1 mile of rail have %.1f%% LOWER ridership\n", 
              abs((exp(coef(lm_binary)[2]) - 1) * 100)))
}
```

---

# Summary: Bus-Rail Interaction Findings

```{r summary-findings}
cat("\n")
cat("=" , rep("=", 70), "\n", sep = "")
cat("SPATIAL ANALYSIS: BUS-RAIL INTERACTION FINDINGS\n")
cat("=" , rep("=", 70), "\n", sep = "")

cat("\n1. DISTANCE TO RAIL METRICS:\n")
cat(sprintf("   - Mean Distance: %.2f miles\n", mean(tract_rail_distance$Dist_to_Rail_mi)))
cat(sprintf("   - Median Distance: %.2f miles\n", median(tract_rail_distance$Dist_to_Rail_mi)))
cat(sprintf("   - Tracts within 1 mile: %.1f%%\n", 
            mean(tract_rail_distance$Near_Rail) * 100))

cat("\n2. RIDERSHIP PATTERNS BY RAIL PROXIMITY:\n")
cat(sprintf("   - Correlation (Distance vs Ridership): %.4f\n", dist_ridership_cor))
cat(sprintf("   - Near Rail Mean: %.1f boardings\n", mean(near_rail_ridership)))
cat(sprintf("   - Far Rail Mean: %.1f boardings\n", mean(far_rail_ridership)))
cat(sprintf("   - Statistical Significance: p = %.4f\n", t_test_result$p.value))

cat("\n3. RECOVERY PATTERNS:\n")
near_recovery <- recovery_by_access %>%
  filter(Rail_Access %in% c("Very Close (≤0.5 mi)", "Close (0.5-1 mi)")) %>%
  pull(Recovery_Ratio) %>%
  mean()
far_recovery <- recovery_by_access %>%
  filter(Rail_Access %in% c("Far (2-5 mi)", "Very Far (>5 mi)")) %>%
  pull(Recovery_Ratio) %>%
  mean()
cat(sprintf("   - Near Rail Recovery: %.1f%%\n", near_recovery * 100))
cat(sprintf("   - Far Rail Recovery: %.1f%%\n", far_recovery * 100))

cat("\n4. REGRESSION RESULTS:\n")
cat(sprintf("   - Distance Coefficient: %.4f\n", coef(lm_county_fe)[2]))
cat(sprintf("   - Within 1 Mile Effect: %.4f log points\n", coef(lm_binary)[2]))
cat(sprintf("   - Model R-squared: %.4f\n", summary(lm_county_fe)$r.squared))

cat("\n5. KEY INTERPRETATION:\n")
if (coef(lm_county_fe)[2] < 0) {
  cat("   ✓ COMPLEMENTARITY: Tracts near rail have higher bus ridership\n")
  cat("   ✓ Feeder Effect: Buses serve as first/last mile connectors\n")
  cat("   → Policy: Enhance bus-rail integration at transfer points\n")
} else {
  cat("   ✓ SUBSTITUTION: Tracts near rail have lower bus ridership\n")
  cat("   ✓ Competition Effect: Rail reduces bus demand in service area\n")
  cat("   → Policy: Target bus expansion to rail-underserved areas\n")
}

cat("\n6. UNDERSERVED TRACTS:\n")
underserved <- tract_rail_distance %>%
  filter(Dist_to_Rail_mi > 2.0)
cat(sprintf("   - Tracts >2 miles from rail: %s (%.1f%%)\n", 
            nrow(underserved),
            nrow(underserved) / nrow(tract_rail_distance) * 100))
cat("   → These tracts are bus-dependent and should be prioritized\n")

cat("\n" , rep("=", 71), "\n", sep = "")
```

## Export Enhanced Dataset

```{r export-enhanced-data}
# Export final dataset with all spatial metrics for modeling.
df_final_with_spatial <- df_clean_with_rail_buffers %>%
  select(
    Census_Tract_ID, County_Name, Year, Day_of_Week,
    On_, Off_,
    Dist_to_Rail_mi, Dist_to_Rail_km,
    Rail_Access, Near_Rail,
    Within_Half_Mile, Within_One_Mile,
    Nearest_Station, Nearest_Line
  )

# Save for modeling phase.
# write_csv(df_final_with_spatial, "data/septa_with_spatial_rail_metrics.csv")

cat("\nFinal Dataset with Spatial Metrics:\n")
cat(sprintf("Observations: %s\n", comma(nrow(df_final_with_spatial))))
cat(sprintf("Variables: %d\n", ncol(df_final_with_spatial)))
cat("\nKey Variables Added:\n")
cat("  - Dist_to_Rail_mi: Continuous distance to nearest station\n")
cat("  - Rail_Access: Categorical proximity classification\n")
cat("  - Near_Rail: Binary indicator (within 1 mile)\n")
cat("  - Within_Half_Mile: Binary buffer indicator\n")
cat("  - Nearest_Station: Name of closest rail station\n")
cat("\nReady for Negative Binomial model with spatial rail features!\n")
```
