---
title: Evictions
subtitle: Exploratory Data Analysis
date: today
author:
  - name: Angel Rutherford
  - name: Ixchel Ramirez
  - name: Tess Vu
    email:
      - tessavu@proton.me
      - tessavu@upenn.edu
affiliation:
  - name: University of Pennsylvania
    department: Urban Spatial Analytics (MUSA)
    city: Philadelphia
    state: PA
    url: https://www.design.upenn.edu/urban-spatial-analytics
format:
  html:
    code-fold: show
    toc: true
    toc_float: true
    toc-expand: true
    smooth-scroll: true
    embed-resources: true
    title-block-style: default
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Potential Project?

**Research Question:** Can we identify tipping point neighborhoods where eviction filings will surge next month to help with preemptive rental assistance distribution?

**Method:** Negative Binomial regression with rolling window cross-validation using monthly tract-level panel data. This could be an issue bc the data is quite zero inflated.

**Data Scope:** January 2020 - November 2025 eviction filings across Philadelphia census tracts with weekly and monthly granularity.

**Key Stuff:**

- Extreme zero-inflation (60-70% zeros) is p bad for Negative Binomial ...
- Strong seasonality with summer-fall peaks requiring month fixed effects.
- Moratorium policy created structural break from March 2020 through September 2021.
- Racial majority strongly predicts filing intensity suggesting equity concerns.
- High-frequency weekly data captures volatility but monthly data better for forecasting.

# Setup and Initial Data Loading

```{r setup}
# Spatial
library(sf)

# Data wrangling
library(tidyverse)
library(lubridate)

# Visualization
library(scales)
library(patchwork)
library(viridis)

# Tables
library(knitr)
library(kableExtra)

options(scipen = 999)

theme_set(
  theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold", size = 14))
)

# Data paths — all sourced from OpenDataPhilly.
path_monthly <- "../data/eviction/philadelphia_monthly_2020_2021.csv"
path_weekly <- "../data/eviction/philadelphia_weekly_2020_2021.csv"
path_claims <- "../data/eviction/philadelphia_claims_monthly.csv"
path_tracts <- "../data/pa_tracts/pa_tracts.shp"

# Policy date boundaries.
date_moratorium_start <- as.Date("2020-03-01")
date_moratorium_end   <- as.Date("2021-09-30")
```

## Monthly Tract-Level Data

```{r load-monthly-data}
df_monthly_raw <- read.csv(path_monthly)

print(dim(df_monthly_raw))
print(names(df_monthly_raw))
```

```{r load-monthly-data-table}
head(df_monthly_raw, 10) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Weekly Tract-Level Data

```{r load-weekly-data}
df_weekly_raw <- read.csv(path_weekly)

print(dim(df_weekly_raw))
```

```{r load-weekly-data-table}
head(df_weekly_raw, 10) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Claims Data

```{r load-claims-data}
df_claims_raw <- read.csv(path_claims)

print(dim(df_claims_raw))
```

```{r load-claims-data-table}
head(df_claims_raw, 10) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Data Cleaning and Feature Engineering

## Monthly Data Preparation

```{r clean-monthly}
df_monthly <- df_monthly_raw %>%
  rename(filings_count = filings_2020) %>%
  mutate(
    date = as.Date(paste0("01/", month), format = "%d/%m/%Y"),
    year = year(date),
    month_num = month(date),
    month_name = month(date, label = TRUE, abbr = FALSE)
  ) %>%
  filter(!is.na(date)) %>%
  mutate(GEOID = as.character(GEOID)) %>%
  # Eviction moratorium indicator and period labels.
  mutate(
    Moratorium_Active = ifelse(
      date >= date_moratorium_start & date <= date_moratorium_end,
      1,
      0
    ),
    Period = case_when(
      date < date_moratorium_start ~ "Pre-Moratorium",
      date >= date_moratorium_start & date <= date_moratorium_end ~ "Moratorium",
      date > date_moratorium_end ~ "Post-Moratorium"
    ),
    Period = factor(Period, levels = c("Pre-Moratorium", "Moratorium", "Post-Moratorium"))
  ) %>%
  mutate(
    Season = case_when(
      month_num %in% c(12, 1, 2)  ~ "Winter",
      month_num %in% c(3, 4, 5)   ~ "Spring",
      month_num %in% c(6, 7, 8)   ~ "Summer",
      month_num %in% c(9, 10, 11) ~ "Fall"
    ),
    Season = factor(Season, levels = c("Winter", "Spring", "Summer", "Fall"))
  )

summary(df_monthly %>%
          select(filings_count, date, year, Moratorium_Active)) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r clean-monthly-coverage}
df_monthly %>%
  summarize(
    Min_Date = min(date),
    Max_Date = max(date),
    N_Months = n_distinct(date),
    N_Tracts = n_distinct(GEOID),
    Total_Obs = n()
  ) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Weekly Data Preparation

```{r clean-weekly}
df_weekly <- df_weekly_raw %>%
  rename(filings_count = filings_2020) %>%
  mutate(
    date = as.Date(week_date),
    year = year(date),
    month_num = month(date),
    week_of_year = week(date)
  ) %>%
  filter(!is.na(date)) %>%
  mutate(
    GEOID = as.character(GEOID),
    Moratorium_Active = ifelse(
      date >= date_moratorium_start & date <= date_moratorium_end,
      1,
      0
    )
  )

df_weekly %>%
  summarize(
    Min_Date = min(date),
    Max_Date = max(date),
    N_Weeks = n_distinct(date),
    N_Tracts = n_distinct(GEOID),
    Total_Obs = n()
  ) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Claims Data Preparation

```{r clean-claims}
df_claims <- df_claims_raw %>%
  mutate(
    date = as.Date(month_date),
    year = year(date),
    month_num = month(date)
  ) %>%
  filter(!is.na(date)) %>%
  mutate(claim_ratio = median_claim / median_claim_baseline)

summary(df_claims %>%
          select(median_claim, median_claim_baseline, claim_ratio)) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Distribution Analysis

## Monthly Filing Distribution

```{r monthly-distribution}
#| fig-dpi: 300
#| fig-height: 10
#| fig-width: 12

zero_stats_monthly <- df_monthly %>%
  summarize(
    Total_Obs = n(),
    Zero_Count = sum(filings_count == 0),
    Zero_Pct = Zero_Count / Total_Obs * 100,
    Positive_Count = sum(filings_count > 0),
    Mean_All = mean(filings_count),
    Mean_Positive = mean(filings_count[filings_count > 0]),
    Median_All = median(filings_count),
    Variance = var(filings_count),
    Dispersion_Ratio = Variance / Mean_All
  )

cat("Zero-Inflation Statistics (Monthly Data):\n")
cat(sprintf("Total Observations: %s\n", comma(zero_stats_monthly$Total_Obs)))
cat(sprintf(
  "Zero Filings: %s (%.1f%%)\n",
  comma(zero_stats_monthly$Zero_Count),
  zero_stats_monthly$Zero_Pct
))
cat(sprintf(
  "Positive Filings: %s (%.1f%%)\n",
  comma(zero_stats_monthly$Positive_Count),
  100 - zero_stats_monthly$Zero_Pct
))
cat(sprintf("\nMean (all): %.2f\n", zero_stats_monthly$Mean_All))
cat(sprintf("Mean (positive only): %.2f\n", zero_stats_monthly$Mean_Positive))
cat(sprintf("Variance: %.2f\n", zero_stats_monthly$Variance))
cat(sprintf("Dispersion Ratio: %.2f\n", zero_stats_monthly$Dispersion_Ratio))

# Raw and log-transformed panels for faceted histogram.
df_hist_monthly <- df_monthly %>%
  mutate(
    Log_Count = log10(filings_count + 1),
    Scale = factor("Raw Count", levels = c("Raw Count", "Log10(Count + 1)"))
  )

df_log_monthly <- df_hist_monthly %>%
  mutate(
    filings_count = Log_Count,
    Scale = factor("Log10(Count + 1)", levels = c("Raw Count", "Log10(Count + 1)"))
  )
```

High zero percentage could use zero-inflated Negative Binomial model, but I think we're restricted to just regular NB, lmk if I'm wrong.
Dispersion ratio > 1 justifies Negative Binomial over Poisson.

```{r monthly-distribution-viz}
df_final_hist_monthly <- bind_rows(df_hist_monthly, df_log_monthly)

dist_plot_monthly <- ggplot(df_final_hist_monthly, aes(x = filings_count)) +
  geom_histogram(bins = 50, fill = "#E74C3C", color = "white", alpha = 0.8) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Distribution of Monthly Eviction Filings per Census Tract",
    subtitle = sprintf(
      "%.1f%% zeros justify Zero-Inflated Negative Binomial (ZINB) model.\nNot sure if in project toolbox.",
      zero_stats_monthly$Zero_Pct
    ),
    x = "Filings Count",
    y = "Frequency (Number of Observations)",
    caption = "Data: Philadelphia Eviction Filings 2020-2025 | Monthly tract-level observations"
  ) +
  facet_wrap(~Scale, scales = "free", ncol = 2) +
  theme(strip.text = element_text(face = "bold"))

dist_plot_monthly
```

## Weekly Filing Distribution

```{r weekly-distribution}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 12

zero_stats_weekly <- df_weekly %>%
  summarize(
    Total_Obs = n(),
    Zero_Count = sum(filings_count == 0),
    Zero_Pct = Zero_Count / Total_Obs * 100,
    Mean_All = mean(filings_count),
    Variance = var(filings_count),
    Dispersion_Ratio = Variance / Mean_All
  )

cat("\nZero-Inflation Statistics (Weekly Data):\n")
cat(sprintf("Total Observations: %s\n", comma(zero_stats_weekly$Total_Obs)))
cat(sprintf(
  "Zero Filings: %s (%.1f%%)\n",
  comma(zero_stats_weekly$Zero_Count),
  zero_stats_weekly$Zero_Pct
))
cat(sprintf("Mean: %.2f\n", zero_stats_weekly$Mean_All))
cat(sprintf("Dispersion Ratio: %.2f\n", zero_stats_weekly$Dispersion_Ratio))
```

```{r weekly-distribution-histogram}
weekly_hist <- df_weekly %>%
  filter(filings_count > 0) %>%
  ggplot(aes(x = filings_count)) +
  geom_histogram(bins = 30, fill = "#3498DB", color = "white", alpha = 0.8) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Distribution of Weekly Eviction Filings (Positive Counts Only)",
    subtitle = sprintf("%.1f%% of observations are zeros (not shown)", zero_stats_weekly$Zero_Pct),
    x = "Weekly Filings Count",
    y = "Frequency",
    caption = "Data: Philadelphia Weekly Eviction Filings | Excludes zero observations"
  )

weekly_hist
```

## Quantile Analysis

```{r quantile-analysis}
quantile_stats_monthly <- df_monthly %>%
  filter(filings_count > 0) %>%
  summarize(
    Min = min(filings_count),
    Q01 = quantile(filings_count, 0.01),
    Q05 = quantile(filings_count, 0.05),
    Q25 = quantile(filings_count, 0.25),
    Median = median(filings_count),
    Q75 = quantile(filings_count, 0.75),
    Q95 = quantile(filings_count, 0.95),
    Q99 = quantile(filings_count, 0.99),
    Max = max(filings_count)
  )

quantile_stats_monthly %>%
  pivot_longer(everything(), names_to = "Quantile", values_to = "Filings") %>%
  kable(digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Temporal Analysis

## Monthly Aggregate Trend with Policy Markers

```{r monthly-system-trend-plot}
#| fig-dpi: 300
#| fig-height: 7
#| fig-width: 14

monthly_aggregate <- df_monthly %>%
  group_by(date, Period) %>%
  summarize(
    Total_Filings = sum(filings_count, na.rm = TRUE),
    Mean_Filings = mean(filings_count, na.rm = TRUE),
    Median_Filings = median(filings_count, na.rm = TRUE),
    N_Tracts = n_distinct(GEOID),
    .groups = "drop"
  )

monthly_trend_plot <- ggplot(monthly_aggregate, aes(x = date, y = Total_Filings)) +
  geom_line(color = "#C0392B", linewidth = 1.3) +
  geom_point(aes(color = Period), size = 3) +
  geom_vline(
    xintercept = date_moratorium_start,
    linetype = "dashed", color = "#27AE60", linewidth = 1
  ) +
  annotate("text",
    x = date_moratorium_start,
    y = max(monthly_aggregate$Total_Filings) * 0.95,
    label = "Moratorium\nStarts", hjust = -0.1, color = "#27AE60", size = 4, fontface = "bold"
  ) +
  geom_vline(
    xintercept = date_moratorium_end,
    linetype = "dashed", color = "#E67E22", linewidth = 1
  ) +
  annotate("text",
    x = date_moratorium_end,
    y = max(monthly_aggregate$Total_Filings) * 0.85,
    label = "Moratorium\nEnds", hjust = 1.1, color = "#E67E22", size = 4, fontface = "bold"
  ) +
  scale_y_continuous(labels = comma) +
  scale_color_manual(values = c(
    "Pre-Moratorium"  = "#3498DB",
    "Moratorium"      = "#27AE60",
    "Post-Moratorium" = "#E67E22"
  )) +
  labs(
    title = "Total Monthly Eviction Filings: System-Wide Trend (2020-2025)",
    subtitle = "Dramatic suppression during moratorium followed by surge after policy ended",
    x = "Date (Month)",
    y = "Total Monthly Filings Across All Census Tracts",
    color = "Policy Period",
    caption = "Data: Philadelphia Eviction Filings | Vertical lines mark policy changes"
  ) +
  theme(legend.position = "bottom")

monthly_trend_plot
```

```{r monthly-system-trend}
monthly_aggregate %>%
  group_by(
    date >= date_moratorium_start & date <= date_moratorium_end
  ) %>%
  summarize(
    Period = ifelse(
      unique(`date >= date_moratorium_start & date <= date_moratorium_end`),
      "Moratorium", "Non-Moratorium"
    ),
    Mean_Monthly = mean(Total_Filings),
    Median_Monthly = median(Total_Filings),
    Min_Monthly = min(Total_Filings),
    Max_Monthly = max(Total_Filings),
    .groups = "drop"
  ) %>%
  select(-1) %>%
  kable(digits = 0) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Weekly Aggregate Trend

```{r weekly-system-trend}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 14

weekly_aggregate <- df_weekly %>%
  group_by(date) %>%
  summarize(
    Total_Filings = sum(filings_count, na.rm = TRUE),
    .groups = "drop"
  )

weekly_trend_plot <- ggplot(weekly_aggregate, aes(x = date, y = Total_Filings)) +
  geom_line(color = "#8E44AD", alpha = 0.6, linewidth = 0.8) +
  geom_smooth(method = "loess", se = TRUE, color = "#E74C3C", linewidth = 1.2, span = 0.1) +
  geom_vline(
    xintercept = date_moratorium_start,
    linetype = "dashed", color = "#27AE60", alpha = 0.7
  ) +
  geom_vline(
    xintercept = date_moratorium_end,
    linetype = "dashed", color = "#E67E22", alpha = 0.7
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Weekly Eviction Filings: High-Frequency Volatility (2020-2025)",
    subtitle = "Smoothed trend (red) shows underlying pattern amid weekly fluctuations",
    x = "Date (Week Ending)",
    y = "Total Weekly Filings Across All Census Tracts",
    caption = "Data: Philadelphia Weekly Eviction Filings | Vertical Lines = Policy Changes"
  )

weekly_trend_plot
```

## Seasonality Analysis

```{r seasonality-analysis-plot}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 12

seasonality_data <- df_monthly %>%
  filter(Period != "Moratorium") %>%
  group_by(month_num, month_name) %>%
  summarize(
    Mean_Filings = mean(filings_count, na.rm = TRUE),
    Median_Filings = median(filings_count, na.rm = TRUE),
    SD_Filings = sd(filings_count, na.rm = TRUE),
    N_Obs = n(),
    .groups = "drop"
  )

seasonality_plot <- ggplot(seasonality_data, aes(x = month_num, y = Mean_Filings)) +
  geom_line(color = "#16A085", linewidth = 1.3) +
  geom_point(color = "#16A085", size = 4) +
  geom_ribbon(
    aes(
      ymin = Mean_Filings - SD_Filings,
      ymax = Mean_Filings + SD_Filings
    ),
    alpha = 0.2, fill = "#16A085"
  ) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(
    title = "Seasonal Pattern of Eviction Filings (Non-Moratorium Periods)",
    subtitle = "Summer-Fall peak suggests back-to-school timing effect | Shaded area = \u00b1 1 SD",
    x = "Month of Year",
    y = "Mean Filings per Tract",
    caption = "Data: Philadelphia Monthly Eviction Filings | Excludes moratorium period"
  )

seasonality_plot
```

```{r seasonality-analysis}
seasonality_data %>%
  select(month_name, Mean_Filings, Median_Filings, SD_Filings) %>%
  kable(digits = 2, col.names = c("Month", "Mean", "Median", "SD")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Claims Severity Over Time

```{r claims-severity-plot}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 12

claims_plot <- ggplot(df_claims, aes(x = date, y = median_claim)) +
  geom_line(color = "#27AE60", linewidth = 1.2) +
  geom_point(color = "#27AE60", size = 2.5) +
  geom_hline(aes(yintercept = median_claim_baseline),
    linetype = "dashed", color = "#E74C3C", linewidth = 1
  ) +
  annotate("text",
    x = min(df_claims$date), y = df_claims$median_claim_baseline[1] * 1.05,
    label = sprintf(
      "Pre-Pandemic Baseline: $%s",
      comma(df_claims$median_claim_baseline[1])
    ),
    hjust = 0, color = "#E74C3C", fontface = "bold"
  ) +
  geom_vline(
    xintercept = date_moratorium_start,
    linetype = "dotted", color = "gray50", alpha = 0.7
  ) +
  geom_vline(
    xintercept = date_moratorium_end,
    linetype = "dotted", color = "gray50", alpha = 0.7
  ) +
  scale_y_continuous(labels = dollar) +
  labs(
    title = "Median Eviction Claim Amount Over Time",
    subtitle = "Claims increased during moratorium as arrears accumulated then stabilized",
    x = "Date (Month)",
    y = "Median Claim Amount (USD)",
    caption = "Data: Philadelphia Eviction Claims | Dashed line = pre-pandemic baseline"
  )

claims_plot
```

```{r claims-severity}
df_claims %>%
  mutate(Period = case_when(
    date < date_moratorium_start ~ "Pre-Moratorium",
    date >= date_moratorium_start & date <= date_moratorium_end ~ "Moratorium",
    date > date_moratorium_end ~ "Post-Moratorium"
  )) %>%
  group_by(Period) %>%
  summarize(
    Mean_Claim = mean(median_claim, na.rm = TRUE),
    Median_Claim = median(median_claim, na.rm = TRUE),
    Ratio_to_Baseline = mean(claim_ratio, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Spatial and Racial Disparity Analysis

## Filings by Racial Majority

```{r racial-disparity-plot}
#| fig-dpi: 300
#| fig-height: 7
#| fig-width: 12

racial_summary <- df_monthly %>%
  filter(!is.na(racial_majority), racial_majority != "") %>%
  group_by(racial_majority) %>%
  summarize(
    Total_Filings = sum(filings_count, na.rm = TRUE),
    Mean_Filings = mean(filings_count, na.rm = TRUE),
    Median_Filings = median(filings_count, na.rm = TRUE),
    N_Tract_Months = n(),
    N_Unique_Tracts = n_distinct(GEOID),
    .groups = "drop"
  ) %>%
  arrange(desc(Total_Filings))

racial_plot <- ggplot(racial_summary, aes(
  x = reorder(racial_majority, Total_Filings),
  y = Total_Filings
)) +
  geom_col(aes(fill = Total_Filings), width = 0.7) +
  coord_flip() +
  scale_fill_gradient(low = "#BDC3C7", high = "#E74C3C", labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Total Eviction Filings by Census Tract Racial Majority",
    subtitle = "Stark disparities indicate structural inequities requiring policy attention",
    x = "Census Tract Racial Majority",
    y = "Total Eviction Filings (2020-2025)",
    fill = "Total\nFilings",
    caption = "Data: Philadelphia Monthly Eviction Filings"
  ) +
  theme(legend.position = "right")

racial_plot
```

```{r racial-disparity}
racial_summary %>%
  kable(digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Temporal Trends by Racial Majority

```{r racial-temporal}
#| fig-dpi: 300
#| fig-height: 7
#| fig-width: 14

racial_temporal <- df_monthly %>%
  filter(!is.na(racial_majority), racial_majority != "") %>%
  group_by(date, racial_majority) %>%
  summarize(Total_Filings = sum(filings_count, na.rm = TRUE), .groups = "drop")

racial_temporal_plot <- ggplot(
  racial_temporal,
  aes(x = date, y = Total_Filings, color = racial_majority)
) +
  geom_line(linewidth = 1.1, alpha = 0.8) +
  geom_vline(
    xintercept = date_moratorium_start,
    linetype = "dashed", color = "gray50", alpha = 0.5
  ) +
  geom_vline(
    xintercept = date_moratorium_end,
    linetype = "dashed", color = "gray50", alpha = 0.5
  ) +
  scale_y_continuous(labels = comma) +
  scale_color_viridis_d(option = "plasma", end = 0.9) +
  labs(
    title = "Monthly Eviction Filings by Racial Majority: Temporal Trends",
    subtitle = "Differential recovery patterns suggest unequal policy impact across communities",
    x = "Date (Month)",
    y = "Total Monthly Filings",
    color = "Racial Majority",
    caption = "Data: Philadelphia Monthly Eviction Filings | Vertical lines = policy changes"
  ) +
  theme(legend.position = "bottom")

racial_temporal_plot
```

## Weekly Filings by Racial Majority (Stacked Area)

```{r racial-weekly-stacked}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 14

weekly_racial <- df_weekly %>%
  filter(!is.na(racial_majority), racial_majority != "") %>%
  group_by(date, racial_majority) %>%
  summarize(Total_Filings = sum(filings_count, na.rm = TRUE), .groups = "drop")

weekly_racial_plot <- ggplot(
  weekly_racial,
  aes(x = date, y = Total_Filings, fill = racial_majority)
) +
  geom_area(alpha = 0.8, color = "white", linewidth = 0.3) +
  scale_y_continuous(labels = comma) +
  scale_fill_viridis_d(option = "turbo", end = 0.9) +
  labs(
    title = "Weekly Eviction Filings by Racial Majority: Stacked Composition",
    subtitle = "Proportional burden shifts over time revealing differential vulnerability",
    x = "Date (Week Ending)",
    y = "Total Weekly Filings",
    fill = "Racial Majority",
    caption = "Data: Philadelphia Weekly Eviction Filings"
  ) +
  theme(legend.position = "bottom")

weekly_racial_plot
```

# Tract-Level Heterogeneity

## High-Risk Tract Identification

```{r high-risk-tracts-table}
tract_totals <- df_monthly %>%
  group_by(GEOID, racial_majority) %>%
  summarize(
    Total_Filings = sum(filings_count, na.rm = TRUE),
    Mean_Monthly = mean(filings_count, na.rm = TRUE),
    Median_Monthly = median(filings_count, na.rm = TRUE),
    SD_Monthly = sd(filings_count, na.rm = TRUE),
    CV = SD_Monthly / Mean_Monthly,
    Months_Observed = n(),
    Months_With_Filings = sum(filings_count > 0),
    .groups = "drop"
  )

top_risk_tracts <- tract_totals %>%
  arrange(desc(Total_Filings)) %>%
  head(20)

top_risk_tracts %>%
  select(GEOID, racial_majority, Total_Filings, Mean_Monthly, CV) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r high-risk-tracts-stats}
total_system_filings <- sum(tract_totals$Total_Filings)
top_20_pct <- sum(top_risk_tracts$Total_Filings) / total_system_filings * 100

cat(sprintf("Concentration: Top 20 tracts account for %.1f%% of all filings\n", top_20_pct))
```

# Pre-Pandemic Baseline Comparison

## Weekly Filings vs Baseline Ratio

```{r weekly-baseline-ratio-plot}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 14

weekly_baseline_ratio <- df_weekly %>%
  group_by(date) %>%
  summarize(
    Total_Filings = sum(filings_count, na.rm = TRUE),
    Total_Baseline = sum(filings_avg_prepandemic_baseline, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(Total_Baseline > 0) %>%
  mutate(Ratio_to_Baseline = Total_Filings / Total_Baseline)

baseline_ratio_plot <- ggplot(weekly_baseline_ratio, aes(x = date, y = Ratio_to_Baseline)) +
  geom_line(color = "#16A085", linewidth = 0.8, alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "#E74C3C", linewidth = 1.2, span = 0.15) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "black", linewidth = 1) +
  annotate("text",
    x = min(weekly_baseline_ratio$date), y = 1.05,
    label = "Pre-Pandemic Average", hjust = 0, fontface = "bold"
  ) +
  geom_vline(
    xintercept = date_moratorium_start,
    linetype = "dotted", color = "gray50", alpha = 0.7
  ) +
  geom_vline(
    xintercept = date_moratorium_end,
    linetype = "dotted", color = "gray50", alpha = 0.7
  ) +
  labs(
    title = "Weekly Filings Normalized to Pre-Pandemic Baseline",
    subtitle = "Ratio > 1 indicates system stress relative to historical norm | Smoothed trend in red",
    x = "Date (Week Ending)",
    y = "Weekly Filings / Pre-Pandemic Average Ratio",
    caption = "Data: Philadelphia Weekly Eviction Filings"
  )

baseline_ratio_plot
```

```{r weekly-baseline-ratio-stats}
weekly_baseline_ratio %>%
  mutate(Period = case_when(
    date < date_moratorium_start ~ "Pre-Moratorium",
    date >= date_moratorium_start & date <= date_moratorium_end ~ "Moratorium",
    date > date_moratorium_end ~ "Post-Moratorium"
  )) %>%
  group_by(Period) %>%
  summarize(
    Mean_Ratio = mean(Ratio_to_Baseline, na.rm = TRUE),
    Median_Ratio = median(Ratio_to_Baseline, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  kable(digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Spatial Analysis

```{r spatial-setup}
tract_geo <- st_read(path_tracts) %>%
  filter(COUNTYFP == "101")

df_monthly_agg <- df_monthly %>%
  group_by(GEOID) %>%
  summarize(
    Total_Filings = sum(filings_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(GEOID = as.character(GEOID))

tract_map_monthly <- tract_geo %>%
  left_join(df_monthly_agg, by = "GEOID")
```

```{r weekly-filings}
df_weekly_agg <- df_weekly %>%
  group_by(GEOID) %>%
  summarize(
    Avg_Weekly_Filings = mean(filings_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(GEOID = as.character(GEOID))

tract_map_weekly <- tract_geo %>%
  left_join(df_weekly_agg, by = "GEOID")
```

## Choropleths

```{r choropleth}
#| fig-width: 12
#| fig-height: 8
#| fig-dpi: 300

common_fill_scale <- scale_fill_viridis_c(
  trans = "log10",
  labels = comma,
  na.value = "gray70",
  direction = -1,
  name = "Count (Log)"
)

map_total_risk <- ggplot(tract_map_monthly) +
  geom_sf(aes(fill = Total_Filings + 1), color = "white", linewidth = 0.1) +
  common_fill_scale +
  labs(
    title = "Total Eviction Filings (2020-2023)",
    subtitle = "Long-term risk and structural disparity."
  ) +
  coord_sf(datum = NA) +
  theme_void()

map_weekly_avg <- ggplot(tract_map_weekly) +
  geom_sf(aes(fill = Avg_Weekly_Filings + 1), color = "white", linewidth = 0.1) +
  common_fill_scale +
  labs(
    title = "Average Weekly Filings",
    subtitle = "Normalized view of short-term filing activity."
  ) +
  coord_sf(datum = NA) +
  theme_void()

(map_total_risk | map_weekly_avg) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom", legend.direction = "horizontal")
```
