---
title: "The Philadelphia Eviction Early Warning System"
subtitle: "A Predictive Model for Proactive Resource Allocation"
date: today
author:
  - name: Angel Rutherford
  - name: Ixchel Ramirez
  - name: Tess Vu
    email:
      - tessavu@proton.me
      - tessavu@upenn.edu
affiliation:
  - name: University of Pennsylvania
    department: Urban Spatial Analytics (MUSA)
    city: Philadelphia
    state: PA
    url: https://www.design.upenn.edu/urban-spatial-analytics
format:
  html:
    code-fold: show
    toc: true
    toc_float: true
    toc-expand: true
    smooth-scroll: true
    embed-resources: true
    title-block-style: default
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# EXECUTIVE SUMMARY

Eviction is both a cause and consequence of poverty that destabilizes entire neighborhoods. Currently, city responses to eviction are reactive, with resources like legal aid and rental assistance deployed *after* filing volumes become a crisis. This project develops a Real-Time Operational Tool for the Philadelphia Office of Homeless Services and the Fair Housing Commission. By shifting from reactive to predictive analysis, we enable the city to allocate limited staff to specific census tracts predicted to experience elevated eviction filings in the coming month.

Our Negative Binomial regression model leverages temporal momentum, spatial spillover effects, policy intervention effects, property tax delinquency stress, and American Community Survey socioeconomic indicators to forecast monthly eviction filing counts at the census tract level.

The model demonstrates strong performance with meaningful improvement, and sets the foundation for building up to a practical and usable tool down the line. Using a robust temporal validation strategy (training through 2023, testing on 2024-2025), the model generalizes well to future periods without overfitting. Also stark racial disparities in eviction burden was identified, with Black-majority tracts accounting for disproportionate shares of filings. These findings emphasize the need for equity-centered implementation safeguards to prevent perpetuating existing disparities through algorithmic resource allocation.

# I. EXPLORATORY DATA ANALYSIS (EDA)

## 1. SETUP AND LIBRARY LOADING

All required R packages were loaded at the start to streamline data cleaning, visualization, spatial analysis, and modeling.

```{r library}
#| message: false
#| warning: false
# Spatial
library(sf)
library(spdep)
library(tidycensus)

# Modeling
library(MASS)
library(broom)

# Data wrangling
library(tidyverse)

# Visualization
library(scales)
library(patchwork)
library(corrplot)
library(knitr)
library(kableExtra)

# Override masking.
select <- dplyr::select
filter <- dplyr::filter

options(scipen = 999)
```

```{r data-paths}
# All input file paths defined here for easy updating.
path_monthly <- "../data/eviction/philadelphia_monthly_2020_2021.csv"
path_weekly <- "../data/eviction/philadelphia_weekly_2020_2021.csv"
path_claims <- "../data/eviction/philadelphia_claims_monthly.csv"
path_tax <- "../data/eviction/real_estate_tax_balances_census_tract.csv"
path_tracts <- "../data/pa_tracts/pa_tracts.shp"
path_neighbors <- "../data/philadelphia-neighborhoods/philadelphia-neighborhoods.shp"
```

Two visualization techniques were employed to provide clear and effective evidence of our findings. First, a consistent plotting theme was established to ensure readability and uniformity across all figures. Second, custom color palettes were applied to highlight key analytical distinctions, specifically racial categories and policy periods. Racial differences in eviction filings are central to the Eviction Lab's dataset and visualizations and were incorporated to acknowledge inequitable outcomes. In addition, the periods before, during, and after the moratorium were visually differentiated to measure the impact of policy interventions on eviction patterns.

```{r customization}
period_colors <- c(
  "Pre-Moratorium" = "#3498DB",
  "Moratorium" = "#27AE60",
  "Post-Moratorium" = "#E67E22"
)

racial_colors <- c(
  "Black" = "#E74C3C",
  "White" = "#3498DB",
  "Hispanic" = "#F39C12",
  "Other" = "#9B59B6"
)

# Shared ggplot theme applied to all figures.
theme_eviction <- function(base_size = 12) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2),
      plot.subtitle = element_text(size = base_size - 1),
      plot.caption = element_text(size = base_size - 3)
    )
}
```

## 2. DATA LOADING AND INITIAL EXPLORATION

### Monthly Tract-Level Eviction Data

The data was accessed and downloaded from the Eviction Lab, an online database that compiles nationwide eviction filings from court records. For states and major cities, the Eviction Lab provides eviction datasets at two temporal resolutions: weekly and monthly. Both scales were initially examined for their distinct analytical value: weekly data highlight short‑term spikes and localized patterns, while monthly data reveal longer‑term trends and align more closely with policy implementation, as months serve as the standard catchment period for resource allocation and housing interventions.

A summary of the dataset's dimensions and preview of the first few rows shows that the monthly eviction data provides space-time insights into filing patterns. The dataset is structured into eight columns:

\begin{enumerate}
  \item Geographic unit of analysis (census tract),
  \item Standard geographic identifier (census tract number),
  \item Majority racial identity of the tract,
  \item Month and year of record,
  \item Eviction filings from 2020--2022, capturing the impact of the COVID-19 pandemic as an extreme destabilizing event and the subsequent moratorium response,
  \item Average filings,
  \item Average filings prior to the pandemic, and
  \item Last upload date of the dataset.
\end{enumerate}

```{r load-monthly-data}
df_monthly_raw <- read_csv(path_monthly, show_col_types = FALSE)

cat("MONTHLY EVICTION DATA DIMENSIONS\n")
cat(sprintf("Rows: %s\n", comma(nrow(df_monthly_raw))))
cat(sprintf("Columns: %d\n", ncol(df_monthly_raw)))
cat(sprintf("Variables: %s\n", paste(names(df_monthly_raw), collapse = ", ")))
```

```{r monthly-data-table}
head(df_monthly_raw, 10) %>%
  kable(
    caption = "Sample of Raw Monthly Eviction Filing Data",
    col.names = c(
      "Type", "GEOID", "Racial  Majority", "Month",
      "Filings", "Average Filings",
      "Pre-Pandemic Average Filings", "Last Updated"
    )
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Weekly Tract-Level Eviction Data

The weekly dataset mirrors the structure of the monthly data set but differs as it contains more rows of data due to its finer granularity and adds additional columns that specific the week relative to the dataset and the calendar date the data was recorded. 

```{r load-weekly-data}
df_weekly_raw <- read_csv(path_weekly, show_col_types = FALSE)

cat("WEEKLY EVICTION DATA DIMENSIONS\n")
cat(sprintf("Rows: %s\n", comma(nrow(df_weekly_raw))))
cat(sprintf("Columns: %d\n", ncol(df_weekly_raw)))
```

```{r weekly-data-table}
head(df_weekly_raw, 10) %>%
  kable(
    caption = "Sample of Raw Weekly Eviction Filing Data",
    col.names = c(
      "Type", "GEOID", "Racial  Majority", "Week", "Date",
      "Filings", "Average Filings",
      "Pre-Pandemic Average Filings", "Last Updated"
    )
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Claims Data

The Eviction Lab provides tract‑level data that summarizes historical and contemporary monthly claims statistics, where claims represent the financial compensation sought by landlords in eviction proceedings. In addition to reporting the median claim amount, the dataset includes thresholds that capture the relative burden these claims pose to renters, offering a spectrum of financial severity across space and time. While these measures are not the primary variables used in this analysis, they provide valuable context for understanding eviction dynamics. Exploratory review suggests that approximately 3% of claims fall below $1,000, which may reflect unpaid fees rather than market‑rate rent, though in some cases it could indicate subsidized public housing rents. Around 6% of claims align with median market rents, while 14% represent unpaid rent for at least six months. These timelines highlight the varied circumstances underlying eviction filings, including the delays introduced by lengthy legal processes. Together, the dataset's structured measures and exploratory insights illustrate both the quantitative severity of claims and the qualitative complexity of eviction proceedings.

```{r load-claims-data}
df_claims_raw <- read_csv(path_claims, show_col_types = FALSE)

cat("CLAIMS DATA DIMENSIONS\n")
cat(sprintf("Rows: %s\n", comma(nrow(df_claims_raw))))
cat(sprintf("Columns: %d\n", ncol(df_claims_raw)))
```

```{r claims-data-table}
head(df_claims_raw, 10) %>%
  kable(
    digits = 2, caption = "Sample of Monthly Claims Severity Data",
    col.names = c(
      "Date", "Median Claim", "Below $1,000",
      "Below Median Rent", "Over Six Months Rent",
      "Baseline Median Claim"
    )
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

This is not used in the dataset, but for exploration behind the scenes of evictions. It looks like 3% of claims are for below $1,000, which could reflect unpaid fees rather than market rate rent. However, it *could* reflect rent in subsidized public housing. 6% is for the median market rate rents. 14% is for rent unpaid for at least 6 months. These timelines could have varying reasons, especially due to the time-consuming legal aspect that could delay filings.

### Real Estate Tax Balances

In our analysis, tax delinquency is treated as a proxy for landlord financial stress; theoretically properties with outstanding tax balances could pressure landlords into pursuing evictions as a means of recovering income or offsetting financial strain. Thus, property tax assessments were explored to determine how financial stress on landlords may shape eviction filings.

```{r load-tax-data}
df_tax_raw <- read_csv(path_tax, show_col_types = FALSE)

cat("TAX DELINQUENCY DATA DIMENSIONS\n")
cat(sprintf("Rows: %s\n", comma(nrow(df_tax_raw))))
cat(sprintf("Columns: %d\n", ncol(df_tax_raw)))
cat(sprintf("Variables: %s\n", paste(names(df_tax_raw), collapse = ", ")))
```

```{r tax-data-table}
head(df_tax_raw, 10) %>%
  kable(
    digits = 2, caption = "Sample of Real Estate Tax Balances by Census Tract",
    col.names = c(
      "Object ID", "Census Tract", "Properties",
      "Minimum Period", "Maximum Period",
      "Principal", "Interest", "Penalty", "Other",
      "Balance", "Average Balance"
    )
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

### Philadelphia Census Tract Geometry

Since all datasets in this analysis are aggregated at the census tract level, a spatial file containing tract boundaries was loaded to access the geographic and geometric attributes of each tract. The inclusion of this file allows for the integration of eviction, claim, property, and other datasets with physical locations for mapping and spatial analysis.

```{r load-spatial-data}
pa_tracts_sf <- st_read(path_tracts, quiet = TRUE)

philly_tracts_sf <- pa_tracts_sf %>%
  filter(str_starts(GEOID, "42101"))

cat("PHILADELPHIA CENSUS TRACT GEOMETRY\n")
cat(sprintf("Total Philadelphia Tracts: %d\n", nrow(philly_tracts_sf)))
cat(sprintf("Coordinate Reference System: %s\n", st_crs(philly_tracts_sf)$input))
```

## 3. DATA CLEANING AND FEATURE ENGINEERING

Certain data cleaning and manipulation techniques were applied across all datasets to establish structural uniformity that would streamline joinability with other datasets. These techniques include removing invalid records, turning geographic identifiers into character strings to ensure they aren't treated as numeric values, and formatting dates. Each dataset was also cleaned and manipulated based on their respective qualities and potential insights.

### Monthly Eviction Data Preparation

To prepare the monthly eviction data for analysis, raw filings were cleaned and enriched with theoretically motivated features. A categorical variable was created to distinguish pre‑moratorium, moratorium, and post‑moratorium periods, alongside a binary flag for months when the moratorium was active. Months were categorized into seasonal indicators as seasonal patterns in filings can be seen in the figures provided on the Eviction Lab's website. 

```{r clean-monthly}
moratorium_start <- as.Date("2020-03-01")
moratorium_end   <- as.Date("2021-09-30")

df_monthly <- df_monthly_raw %>%
  rename(filings_count = filings_2020) %>%
  mutate(
    date = as.Date(paste0("01/", month), format = "%d/%m/%Y"),
    year = year(date),
    month_num = month(date),
    month_name = month(date, label = TRUE, abbr = FALSE)
  ) %>%
  filter(!is.na(date)) %>%
  mutate(
    GEOID = as.character(GEOID),
    moratorium_active = if_else(
      date >= moratorium_start & date <= moratorium_end, 1, 0
    ),
    period = case_when(
      date < moratorium_start  ~ "Pre-Moratorium",
      date <= moratorium_end   ~ "Moratorium",
      date > moratorium_end    ~ "Post-Moratorium"
    ),
    period = factor(period, levels = c("Pre-Moratorium", "Moratorium", "Post-Moratorium")),
    season = case_when(
      month_num %in% c(12, 1, 2)  ~ "Winter",
      month_num %in% c(3, 4, 5)   ~ "Spring",
      month_num %in% c(6, 7, 8)   ~ "Summer",
      month_num %in% c(9, 10, 11) ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    # Capture post-moratorium rebound effect.
    post_moratorium_months = if_else(
      date > moratorium_end,
      as.numeric(difftime(date, moratorium_end, units = "days")) / 30,
      0
    )
  )

cat("CLEANED MONTHLY DATA SUMMARY\n")
summary(df_monthly %>% select(filings_count, date, year, moratorium_active))
```

To verify the scope of the cleaned dataset, a summary table was generated showing the earliest and latest months covered, the total number of distinct months, the number of census tracts represented, and the overall number of observations. This check confirms that the dataset ranges from January 2020 to November 2025.

```{r clean-monthly-coverage}
df_monthly %>%
  summarize(
    min_date = min(date),
    max_date = max(date),
    n_months = n_distinct(date),
    n_tracts = n_distinct(GEOID),
    total_obs = n()
  ) %>%
  kable(
    caption = "Monthly Data Temporal Coverage",
    col.names = c("Minimum Date", "Maximum Date", "Months", "Tracts", "Observations")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Weekly Data Preparation

Weekly data was cleaned and enriched in parallel with the monthly dataset, but adapted to reflect the finer weekly temporal scale. A summary table of temporal coverage reveals that the dataset spans from the final days of December 2019 through the second week of November 2025, offering a more nuanced timeline down to the exact day.

```{r clean-weekly}
df_weekly <- df_weekly_raw %>%
  rename(filings_count = filings_2020) %>%
  mutate(
    date = as.Date(week_date),
    year = year(date),
    month_num = month(date),
    week_of_year = week(date)
  ) %>%
  filter(!is.na(date)) %>%
  mutate(
    GEOID = as.character(GEOID),
    moratorium_active = if_else(
      date >= moratorium_start & date <= moratorium_end, 1, 0
    )
  )

cat("CLEANED WEEKLY DATA SUMMARY\n")
summary(df_weekly %>% select(filings_count, date, year, moratorium_active))
```

```{r clean-weekly-coverage}
df_weekly %>%
  summarize(
    min_date = min(date),
    max_date = max(date),
    n_weeks = n_distinct(date),
    n_tracts = n_distinct(GEOID),
    total_obs = n()
  ) %>%
  kable(
    caption = "Weekly Data Temporal Coverage",
    col.names = c("Minimum Date", "Maximum Date", "Weeks", "Tracts", "Observations")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Claims Data Preparation

Monthly eviction claims data was mutated to include a new column, claims ratio, that was calculated to represent median claim values relative to pre‑pandemic baselines. 

```{r clean-claims}
df_claims <- df_claims_raw %>%
  mutate(
    date = as.Date(month_date),
    year = year(date),
    month_num = month(date)
  ) %>%
  filter(!is.na(date)) %>%
  mutate(claim_ratio = median_claim / median_claim_baseline)

cat("CLAIMS DATA SUMMARY\n")
summary(df_claims %>% select(median_claim, median_claim_baseline, claim_ratio))
```

### Tax Delinquency Data Preparation

We defined delinquency in the property tax dataset as properties that received penalties rather than properties that just have overdue balances. Due to the marginal frequency observed in the dataset's summary statistics (only 8 properties identified as overdue with no penalties), overdue properties were removed. For each tract, counts of delinquent properties, total balances, and average balances per property were calculated, along with log transformations of balances and delinquency age in years. Summary statistics were generated to report mean and median property counts, balances, and average balances across tracts, providing a tract‑level profile of tax delinquency severity.

```{r clean-tax-data}
df_tax <- df_tax_raw %>%
  mutate(census_tract = as.character(census_tract)) %>%
  filter(!str_detect(census_tract, "Other/Unidentified")) %>%
  mutate(
    has_penalty = penalty > 0,
    delinquent_prop_count = if_else(has_penalty, num_props, 0L),
    delinquent_balance = if_else(has_penalty, balance, 0),
    avg_delinquent_balance = if_else(has_penalty, balance / num_props, 0),
    overdue_only_props = if_else(!has_penalty, num_props, 0L),
    overdue_only_balance = if_else(!has_penalty, balance, 0),
    log_delinquent_balance = log1p(delinquent_balance),
    log_overdue_balance = log1p(overdue_only_balance)
  ) %>%
  rename(GEOID = census_tract)

cat("TAX DELINQUENCY DATA SUMMARY\n")
cat(sprintf("Tracts with Tax Data: %d\n", nrow(df_tax)))
cat(sprintf("Total Properties with Any Tax Balance: %s\n", comma(sum(df_tax$num_props))))
cat(sprintf(
  "Delinquent Properties (Penalty > 0): %s (%.1f%%)\n",
  comma(sum(df_tax$delinquent_prop_count)),
  sum(df_tax$delinquent_prop_count) / sum(df_tax$num_props) * 100
))
cat(sprintf(
  "Overdue Properties (No Penalty): %s (%.1f%%)\n",
  comma(sum(df_tax$overdue_only_props)),
  sum(df_tax$overdue_only_props) / sum(df_tax$num_props) * 100
))
cat(sprintf("Total Delinquent Balance: $%s\n", comma(sum(df_tax$delinquent_balance))))
cat(sprintf("Total Overdue-Only Balance: $%s\n", comma(sum(df_tax$overdue_only_balance))))
```

```{r overdue-removal}
# 8 overdue properties is marginal; keep only penalty-bearing records.
df_tax <- df_tax_raw %>%
  mutate(census_tract = as.character(census_tract)) %>%
  filter(
    !str_detect(census_tract, "Other/Unidentified"),
    penalty > 0
  ) %>%
  mutate(
    delinquent_prop_count = num_props,
    delinquent_balance = balance,
    avg_delinquent_balance_per_prop = balance / num_props,
    log_delinquent_balance = log1p(balance),
    delinquency_age_years = 2025 - min_period,
    avg_balance = balance / num_props
  ) %>%
  rename(GEOID = census_tract)
```

```{r tax-summary-stats}
tax_summary <- df_tax %>%
  summarize(
    mean_props = mean(num_props),
    median_props = median(num_props),
    mean_balance = mean(balance),
    median_balance = median(balance),
    mean_avg_balance = mean(avg_balance),
    median_avg_balance = median(avg_balance)
  )

tax_summary %>%
  pivot_longer(everything(), names_to = "statistic", values_to = "value") %>%
  mutate(value = if_else(
    str_detect(statistic, "balance"),
    paste0("$", comma(round(value, 2))),
    comma(round(value, 2))
  )) %>%
  kable(
    caption = "Tax Delinquency Summary Statistics",
    col.names = c("Statistic", "Value")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Sealed Tract Assessment

Monthly eviction filings were assessed for sealed census tracts, where GEOIDs are suppressed for privacy protection. These tracts accounted for a small share of observations but exhibited unusually high filing counts. Because sealed tracts cannot be geographically mapped, they were excluded from spatial modeling to preserve tract‑level specificity. While this exclusion reduces the ability to design geographically actionable policies and outreach, the sealed tracts nonetheless represent legitimate evictions and highlight systematic vulnerabilities that cannot be directly addressed in this analysis.

```{r sealed-tracts}
sealed_count <- df_monthly %>%
  filter(GEOID == "sealed" | is.na(GEOID) | GEOID == "") %>%
  nrow()

sealed_pct <- sealed_count / nrow(df_monthly) * 100

cat("SEALED TRACT ASSESSMENT\n")
cat(sprintf("Sealed or Missing Tract Observations: %s (%.1f%%)\n", comma(sealed_count), sealed_pct))
```

```{r sealed-tracts-stats}
sealed_tracts_analysis <- df_monthly %>%
  filter(GEOID == "sealed" | is.na(GEOID) | GEOID == "") %>%
  summarize(
    n_obs = n(),
    n_unique_dates = n_distinct(date),
    total_filings = sum(filings_count, na.rm = TRUE),
    mean_filings = mean(filings_count, na.rm = TRUE),
    median_filings = median(filings_count, na.rm = TRUE),
    min_filings = min(filings_count, na.rm = TRUE),
    max_filings = max(filings_count, na.rm = TRUE),
    sd_filings = sd(filings_count, na.rm = TRUE),
    zero_pct = sum(filings_count == 0) / n() * 100
  )

cat(sprintf("Total Observations: %s\n", comma(sealed_tracts_analysis$n_obs)))
cat(sprintf("Months of Data: %d\n", sealed_tracts_analysis$n_unique_dates))
cat(sprintf("Total Filings in Sealed Tracts: %s\n", comma(sealed_tracts_analysis$total_filings)))
cat(sprintf("Mean Filings per Tract-Month: %.1f\n", sealed_tracts_analysis$mean_filings))
cat(sprintf("Median Filings per Tract-Month: %.1f\n", sealed_tracts_analysis$median_filings))
cat(sprintf("Range: %d to %d\n", sealed_tracts_analysis$min_filings, sealed_tracts_analysis$max_filings))
cat(sprintf("Standard Deviation: %.1f\n", sealed_tracts_analysis$sd_filings))
cat(sprintf("Zero-Filing Months: %.1f%%\n", sealed_tracts_analysis$zero_pct))
```

```{r sealed-tracts-distribution}
sealed_distribution <- df_monthly %>%
  filter(GEOID == "sealed") %>%
  group_by(filings_count) %>%
  summarize(n = n(), .groups = "drop") %>%
  arrange(desc(filings_count))

head(sealed_distribution, 10) %>%
  kable(
    caption = "Highest Filing Counts in Sealed Tracts",
    col.names = c("Filings", "Frequency")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The sealed tracts are outliers and are legitimate evictions that reflect the current systematic issues, and could represent vulnerable communities experiencing a recent mass eviction (highest unsealed filings are from 2025). They aren't included because the tracts cannot be mapped to in the data if they're sealed, this privacy protection prevents geographically actionable policies and outreach, however, it has to be stressed that this is absolutely *not* ideal.

## 4. EXPLORATORY DATA ANALYSIS

### Distribution Analysis: Zero-Inflation Assessment

Understanding the distribution of eviction filings is critical for selecting the appropriate modeling approach. Because eviction filings are discrete, non‑continuous count values, they require models such as Poisson and Negative Binomial regression specifically designed to handle count data. A major difference between the two approaches is their ability to handle overdispersion, with the Poisson model assuming equal mean and variance while the Negative Binomial model allows the variance to exceed the mean, making it more flexible for data with excess variability and zeros. Zero-inflation statistics were calculated and visualized to assess the distribution of monthly eviction filings. The summary statistics report reveals substantial zero-inflation and variance that exceeded the mean, yielding a dispersion ratio well over 1, indicating that the negative binomial model as a more appropriate modeling approach for our data.

A histogram of raw counts of evictions further highlights the prevalence of zeros and the data's skewed distribution, while a log‑transformed histogram provides a clearer view of variation. 

```{r monthly-distribution}
zero_stats_monthly <- df_monthly %>%
  summarize(
    total_obs = n(),
    zero_count = sum(filings_count == 0),
    zero_pct = zero_count / total_obs * 100,
    positive_count = sum(filings_count > 0),
    mean_all = mean(filings_count),
    mean_positive = mean(filings_count[filings_count > 0]),
    median_all = median(filings_count),
    variance = var(filings_count),
    dispersion_ratio = variance / mean_all
  )

cat("MONTHLY ZERO-INFLATION STATS\n")
cat(sprintf("Total Observations: %s\n", comma(zero_stats_monthly$total_obs)))
cat(sprintf(
  "Zero Filings: %s (%.1f%%)\n",
  comma(zero_stats_monthly$zero_count),
  zero_stats_monthly$zero_pct
))
cat(sprintf(
  "Positive Filings: %s (%.1f%%)\n",
  comma(zero_stats_monthly$positive_count),
  100 - zero_stats_monthly$zero_pct
))
cat(sprintf("Mean (all observations): %.2f\n", zero_stats_monthly$mean_all))
cat(sprintf("Mean (positive only): %.2f\n", zero_stats_monthly$mean_positive))
cat(sprintf("Variance: %.2f\n", zero_stats_monthly$variance))
cat(sprintf("Dispersion Ratio (Variance/Mean): %.2f\n", zero_stats_monthly$dispersion_ratio))
```

```{r monthly-distribution-viz}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 12

monthly_plot_raw <- df_monthly %>%
  ggplot(aes(x = filings_count)) +
  geom_histogram(bins = 30, fill = "#E74C3C", color = "white") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Raw Count Distribution", x = "Monthly Filings Count", y = "Frequency") +
  theme_eviction() +
  theme(plot.title = element_text(hjust = 0.5))

monthly_plot_log <- df_monthly %>%
  mutate(log_filings = log10(filings_count + 1)) %>%
  ggplot(aes(x = log_filings)) +
  geom_histogram(bins = 30, fill = "#E67E22", color = "white") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Log-Transformed Distribution", x = "log(Monthly Filings + 1)", y = "Frequency") +
  theme_eviction() +
  theme(plot.title = element_text(hjust = 0.5))

monthly_plot_raw + monthly_plot_log +
  plot_annotation(
    title = "Distribution of Monthly Eviction Filings by Census Tract",
    subtitle = sprintf(
      "%.1f%% Zeros | Dispersion Ratio = %.1f",
      zero_stats_monthly$zero_pct,
      zero_stats_monthly$dispersion_ratio
    ),
    caption = "Data: Philadelphia Eviction Filings 2020-2025 via Eviction Lab",
    theme = theme_eviction()
  )
```

### Weekly Filing Distribution

Weekly eviction filings were also assessed for zero-inflation and overdispersion; the distribution was highly zero-inflated, making even standard Negative Binomial models unsuitable. However, the data structure is compatible with Zero-Inflated Negative Binomial models, which may be explored in future work. In this analysis we chose to proceed with the monthly eviction data to preserve the performance of the Negative Binomial model.

```{r weekly-distribution}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 12

zero_stats_weekly <- df_weekly %>%
  summarize(
    total_obs = n(),
    zero_count = sum(filings_count == 0),
    zero_pct = zero_count / total_obs * 100,
    mean_all = mean(filings_count),
    variance = var(filings_count),
    dispersion_ratio = variance / mean_all
  )

cat("WEEKLY ZERO-INFLATION STATS\n")
cat(sprintf("Total Observations: %s\n", comma(zero_stats_weekly$total_obs)))
cat(sprintf(
  "Zero Filings: %s (%.1f%%)\n",
  comma(zero_stats_weekly$zero_count),
  zero_stats_weekly$zero_pct
))
cat(sprintf("Mean: %.2f\n", zero_stats_weekly$mean_all))
cat(sprintf("Dispersion Ratio: %.2f\n", zero_stats_weekly$dispersion_ratio))
```

```{r weekly-distribution-histogram}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 12

plot_weekly_raw <- df_weekly %>%
  ggplot(aes(x = filings_count)) +
  geom_histogram(bins = 30, fill = "#3498DB", color = "white") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Raw Count Distribution", x = "Weekly Filings Count", y = "Frequency") +
  theme_eviction() +
  theme(plot.title = element_text(hjust = 0.5))

plot_weekly_log <- df_weekly %>%
  mutate(log_filings = log10(filings_count + 1)) %>%
  ggplot(aes(x = log_filings)) +
  geom_histogram(bins = 30, fill = "#8E44AD", color = "white") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(title = "Log-Transformed Distribution", x = "log(Weekly Filings + 1)", y = "Frequency") +
  theme_eviction() +
  theme(plot.title = element_text(hjust = 0.5))

plot_weekly_raw + plot_weekly_log +
  plot_annotation(
    title = "Distribution of Weekly Eviction Filings by Census Tract",
    subtitle = sprintf(
      "%.1f%% Zeros | Dispersion Ratio = %.1f",
      zero_stats_weekly$zero_pct,
      zero_stats_weekly$dispersion_ratio
    ),
    caption = "Data: Philadelphia Eviction Filings via Eviction Lab",
    theme = theme_eviction()
  )
```

### Tax Delinquency Distribution

The histogram of average delinquent balances per property across census tracts is highly right-skewed, with most tracts showing low balances and a few exhibiting extreme outliers revealing that a small number of tracts carry disproportionately high tax burdens. A histogram of  log transformed tax balances normalizes this skew, producing a more symmetric, bell-shaped distribution suitable for statistical modeling. 

```{r tax-distribution-histogram}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 12

plot_tax_raw <- df_tax %>%
  ggplot(aes(x = avg_delinquent_balance_per_prop)) +
  geom_histogram(bins = 40, fill = "#9B59B6", color = "white") +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Raw Average Balance",
    x = "Average Delinquent Balance per Property ($)",
    y = "Number of Census Tracts"
  ) +
  theme_eviction() +
  theme(plot.title = element_text(hjust = 0.5))

plot_tax_log <- df_tax %>%
  mutate(log_balance = log10(avg_delinquent_balance_per_prop + 1)) %>%
  ggplot(aes(x = log_balance)) +
  geom_histogram(bins = 40, fill = "#3498DB", color = "white") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Log-Transformed",
    x = "log(Average Balance + 1)",
    y = "Number of Census Tracts"
  ) +
  theme_eviction() +
  theme(plot.title = element_text(hjust = 0.5))

plot_tax_raw + plot_tax_log +
  plot_annotation(
    title = "Distribution of Average Tax Delinquency Balance by Census Tract",
    caption = "Data: OpenDataPhilly Real Estate Tax Balances",
    theme = theme_eviction()
  )
```

### Quantile Analysis

To further assess the tail-behavior of monthly eviction filings, statistice for quantiles thresholds were calculated. This included the minimum, 1st, 5th, 25th, 50th (median), 75th, 95th, and 99th percentiles, as well as the maximum. These statistics provide a clear view of the heavy right tail of the distribution, complementing the zero‑inflation and over‑dispersion diagnostics as well as indicating the need to create a variable the addresses abnormal spikes.

```{r quantile-analysis}
quantile_stats_monthly <- df_monthly %>%
  filter(filings_count > 0) %>%
  summarize(
    min = min(filings_count),
    q01 = quantile(filings_count, 0.01),
    q05 = quantile(filings_count, 0.05),
    q25 = quantile(filings_count, 0.25),
    median = median(filings_count),
    q75 = quantile(filings_count, 0.75),
    q95 = quantile(filings_count, 0.95),
    q99 = quantile(filings_count, 0.99),
    max = max(filings_count)
  )

quantile_stats_monthly %>%
  pivot_longer(everything(), names_to = "quantile", values_to = "filings") %>%
  kable(
    digits = 1, caption = "Monthly Filing Count Quantiles (Positive Counts Only)",
    col.names = c("Quantile", "Filings")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 5. TEMPORAL ANALYSIS: SYSTEM-WIDE TRENDS

### Monthly Aggregate Trend with Policy Markers

The plot and summary statistics of monthly evictions overtime show a clear policy-linked trajectory from 2020 to 2025 as filings dropped sharply at the onset of the eviction moratorium, remained suppressed throughout its duration, and rose after the moratorium ended. This pattern underscores the impact of policy interventions as well as highlights the need for temporal controls in modeling.

```{r monthly-system-trend-plot}
#| fig-dpi: 300
#| fig-height: 7
#| fig-width: 14

monthly_aggregate <- df_monthly %>%
  group_by(date, period) %>%
  summarize(
    total_filings = sum(filings_count, na.rm = TRUE),
    mean_filings = mean(filings_count, na.rm = TRUE),
    median_filings = median(filings_count, na.rm = TRUE),
    n_tracts = n_distinct(GEOID),
    .groups = "drop"
  )

monthly_trend_plot <- ggplot(monthly_aggregate, aes(x = date, y = total_filings)) +
  geom_line(color = "#C0392B", linewidth = 1.5) +
  geom_point(aes(color = period), size = 4) +
  geom_vline(
    xintercept = moratorium_start,
    linetype = "dashed", color = "#27AE60", linewidth = 1
  ) +
  annotate("text",
    x = moratorium_start, y = max(monthly_aggregate$total_filings) * 0.95,
    label = "Moratorium\nStarts", hjust = -0.1, color = "#27AE60", size = 4, fontface = "bold"
  ) +
  geom_vline(
    xintercept = moratorium_end,
    linetype = "dashed", color = "#E67E22", linewidth = 1
  ) +
  annotate("text",
    x = moratorium_end, y = max(monthly_aggregate$total_filings) * 0.85,
    label = "Moratorium\nEnds", hjust = 1.1, color = "#E67E22", size = 4, fontface = "bold"
  ) +
  scale_y_continuous(labels = comma) +
  scale_color_manual(values = period_colors) +
  labs(
    title = "Total Monthly Eviction Filings: System-Wide Trend (2020-2025)",
    subtitle = "Suppression during moratorium followed by sustained elevation after end.",
    x = "Date (Month)",
    y = "Total Monthly Filings Across All Census Tracts",
    color = "Policy Period",
    caption = "Data: Philadelphia Eviction Filings via Eviction Lab"
  ) +
  theme_eviction() +
  theme(legend.position = "bottom")

monthly_trend_plot
```

```{r monthly-system-trend-table}
monthly_aggregate %>%
  group_by(period) %>%
  summarize(
    n_months = n(),
    mean_monthly = mean(total_filings),
    median_monthly = median(total_filings),
    min_monthly = min(total_filings),
    max_monthly = max(total_filings),
    sd_monthly = sd(total_filings),
    .groups = "drop"
  ) %>%
  kable(
    digits = 0, caption = "Monthly Filing Statistics by Policy Period",
    col.names = c(
      "Period", "Months", "Monthly Mean", "Monthly Median",
      "Monthly Minimum", "Monthly Maximum", "Monthly Standard Deviation"
    )
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Seasonality Analysis

The box plot of post moratorium evictions by season and month show elevated filings in winter and summer as well as monthly alternating patterns of elevation in the winter to early spring months. Summary statistics of both plots illustrates that evictions typically remained stable across seasons and months but do possess slightly higher numbers consistent with the visual patterns.

```{r seasonality-boxplot}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 14

seasonality_data <- df_monthly %>%
  filter(period == "Post-Moratorium") %>%
  group_by(season) %>%
  summarize(
    mean_filings = mean(filings_count, na.rm = TRUE),
    median_filings = median(filings_count, na.rm = TRUE),
    sd_filings = sd(filings_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(season)

seasonality_boxplot <- seasonality_data %>%
  ggplot(aes(x = season, y = mean_filings)) +
  geom_col(fill = "#3498DB") +
  geom_errorbar(
    aes(ymin = mean_filings - sd_filings, ymax = mean_filings + sd_filings),
    width = 0.3
  ) +
  labs(title = "Seasonal Pattern in Eviction Filings", x = "Season", y = "Mean Filings per Tract-Month") +
  theme_eviction() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

monthly_data <- df_monthly %>%
  filter(period == "Post-Moratorium") %>%
  group_by(month_name, month_num) %>%
  summarize(
    mean_filings = mean(filings_count, na.rm = TRUE),
    median_filings = median(filings_count, na.rm = TRUE),
    sd_filings = sd(filings_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(month_num)

monthly_boxplot <- monthly_data %>%
  ggplot(aes(x = reorder(month_name, month_num), y = mean_filings)) +
  geom_col(fill = "#3498DB") +
  geom_errorbar(
    aes(ymin = mean_filings - sd_filings, ymax = mean_filings + sd_filings),
    width = 0.3
  ) +
  labs(title = "Monthly Pattern in Eviction Filings", x = "Month", y = "Mean Filings per Tract-Month") +
  theme_eviction() +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

seasonality_boxplot + monthly_boxplot +
  plot_annotation(
    title = "Seasonal and Monthly Patterns in Eviction Filings",
    subtitle = "Post-Moratorium",
    caption = "Data: Philadelphia Eviction Claims via Eviction Lab",
    theme = theme_eviction()
  )
```

```{r seasonality-table}
seasonality_data %>%
  kable(
    digits = 2, caption = "Seasonal Stats (Post-Moratorium Period)",
    col.names = c("Season", "Mean Filings", "Median Filings", "Standard Deviation Filings")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r monthly-table}
monthly_data %>%
  kable(
    digits = 2, caption = "Monthly Stats (Post-Moratorium Period)",
    col.names = c("Month", "Month Number", "Mean Filings", "Median Filings", "Standard Deviation Filings")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Claims Severity Over Time

A temporal plot shows that median eviction claim amounts rose sharply at the onset of the pandemic, peaking in 2021–2022 before gradually declining and stabilizing near the pre-pandemic baseline, suggesting a temporary inflation in claim severity during the moratorium and immediate post-moratorium periods.

```{r claims-trend}
#| fig-dpi: 300
#| fig-height: 7
#| fig-width: 14

claims_plot <- df_claims %>%
  filter(!is.na(median_claim)) %>%
  ggplot(aes(x = date, y = median_claim)) +
  geom_line(color = "#27AE60", linewidth = 1.5) +
  geom_point(color = "#27AE60", size = 3) +
  geom_hline(aes(yintercept = median_claim_baseline),
    linetype = "dashed", color = "#E74C3C", linewidth = 1
  ) +
  annotate("text",
    x = min(df_claims$date - 100), y = df_claims$median_claim_baseline[1] + 200,
    label = sprintf("Pre-Pandemic Baseline: $%s", comma(df_claims$median_claim_baseline[1])),
    hjust = 0, color = "#E74C3C", fontface = "bold"
  ) +
  geom_vline(xintercept = moratorium_start, linetype = "dotted", color = "#27AE60", linewidth = 1) +
  geom_vline(xintercept = moratorium_end, linetype = "dotted", color = "#E67E22", linewidth = 1) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title = "Median Eviction Claim Amount Over Time",
    x = "Date (Month)",
    y = "Median Claim Amount ($)",
    caption = "Data: Philadelphia Eviction Claims via Eviction Lab"
  ) +
  theme_eviction()

claims_plot
```

## 6. RACIAL DISPARITY ANALYSIS

### Filing Burden by Racial Majority

Summary statistics and plots reveal stark disparities across tract racial majorities:  51.9% of filings occurred in Black-majority tracts, despite these tracts representing a smaller share of the city overall. In contrast, White-majority tracts accounted for just 22.1% of filings, though they comprised 36.0% of the comparison group. Hispanic-majority tracts showed near parity, while "Other" tracts had slightly lower filing rates than their representation. These patterns suggest disproportionate eviction exposure in Black communities, underscoring the need for equity-based analysis of our model's performance.

```{r racial-disparity}
#| fig-dpi: 300
#| fig-height: 8
#| fig-width: 12

racial_stats <- df_monthly %>%
  filter(!is.na(racial_majority), racial_majority != "") %>%
  group_by(racial_majority) %>%
  summarize(
    n_obs = n(),
    n_tracts = n_distinct(GEOID),
    total_filings = sum(filings_count, na.rm = TRUE),
    mean_filings = mean(filings_count, na.rm = TRUE),
    median_filings = median(filings_count, na.rm = TRUE),
    sd_filings = sd(filings_count, na.rm = TRUE),
    zero_pct = sum(filings_count == 0) / n() * 100,
    .groups = "drop"
  ) %>%
  mutate(
    pct_total_filings = total_filings / sum(total_filings) * 100,
    pct_tracts = n_tracts / sum(n_tracts) * 100
  )

racial_stats %>%
  kable(
    digits = 2, caption = "Eviction Filing Statistics by Tract Racial Majority",
    col.names = c(
      "Racial Majority", "Observations", "Tracts", "Filings",
      "Mean Filings", "Median Filings", "Standard Deviation Filings",
      "% Zero", "% Total Filings", "% Tracts"
    )
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r racial-disparity-plot}
#| fig-dpi: 300
#| fig-height: 8
#| fig-width: 12

racial_comparison <- racial_stats %>%
  select(racial_majority, pct_tracts, pct_total_filings) %>%
  pivot_longer(
    cols = c(pct_tracts, pct_total_filings),
    names_to = "metric", values_to = "percentage"
  ) %>%
  mutate(metric = if_else(metric == "pct_tracts", "Share of Tracts", "Share of Filings"))

racial_disparity_plot <- racial_comparison %>%
  ggplot(aes(x = racial_majority, y = percentage, fill = metric)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", percentage)),
    position = position_dodge(width = 0.7), vjust = -0.5, size = 3.5
  ) +
  scale_fill_manual(values = c("Share of Tracts" = "#3498DB", "Share of Filings" = "#E74C3C")) +
  scale_y_continuous(labels = percent_format(scale = 1), expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Racial Disparity in Eviction Filings",
    x = "Tract Racial Majority",
    y = "Percentage",
    fill = "Metric",
    caption = "Data: Philadelphia Eviction Filings via Eviction Lab"
  ) +
  theme_eviction() +
  theme(legend.position = "bottom")

racial_disparity_plot
```

## 7. GEOGRAPHIC ANALYSIS

### High-Risk Tract Identification

High‑risk tracts were identified by calculating cumulative eviction filings across the study period to pinpoint neighborhoods with disproportionate concentrations of cases. The analysis shows that the top 20 census tracts alone accounted for 16.6% of all filings.

```{r high-risk-tracts}
tract_totals <- df_monthly %>%
  filter(GEOID != "sealed", !is.na(GEOID), GEOID != "") %>%
  group_by(GEOID, racial_majority) %>%
  summarize(
    total_filings = sum(filings_count, na.rm = TRUE),
    mean_monthly = mean(filings_count, na.rm = TRUE),
    median_monthly = median(filings_count, na.rm = TRUE),
    sd_monthly = sd(filings_count, na.rm = TRUE),
    # Coefficient of variation: standardized dispersion. Higher = more volatile.
    cv = if_else(mean_monthly > 0, sd_monthly / mean_monthly, NA_real_),
    months_observed = n(),
    months_with_filings = sum(filings_count > 0),
    .groups = "drop"
  )

top_risk_tracts <- tract_totals %>%
  arrange(desc(total_filings)) %>%
  head(20)

top_risk_tracts %>%
  select(GEOID, racial_majority, total_filings, mean_monthly, cv) %>%
  kable(
    digits = 2, caption = "Top 20 Highest-Risk Census Tracts (Total Filings 2020-2025)",
    col.names = c("GEOID", "Racial Majority", "Filings", "Monthly Mean", "Coefficient of Variation")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r high-risk-tracts-stats}
total_system_filings <- sum(tract_totals$total_filings)
top_20_pct <- sum(top_risk_tracts$total_filings) / total_system_filings * 100

cat("GEOGRAPHIC CONCENTRATION ANALYSIS\n")
cat(sprintf("Total System Filings: %s\n", comma(total_system_filings)))
cat(sprintf("Top 20 Tracts Account For: %.1f%% of all filings\n", top_20_pct))
```

### Spatial Mapping of Eviction Burden

The choropleth plot showing the spatial distribution of evictions reveals regional concentration of census tracts with high filing levels in North, Upper North, and West Philadelphia.

```{r spatial-eviction-map}
#| fig-dpi: 300
#| fig-height: 10
#| fig-width: 10

philly_eviction_sf <- philly_tracts_sf %>%
  left_join(tract_totals, by = "GEOID")

eviction_map <- ggplot(philly_eviction_sf) +
  geom_sf(aes(fill = total_filings), color = NA) +
  scale_fill_viridis_c(
    option = "inferno",
    trans = "log1p",
    labels = comma,
    na.value = "gray50",
    breaks = c(0, 25, 100, 250, 831)
  ) +
  labs(
    title = "Total Eviction Filings by Census Tract (2020-2025)",
    fill = "Total\nFilings",
    caption = "Data: Philadelphia Eviction Filings via Eviction Lab"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "right"
  )

eviction_map
```

### Spatial Mapping of Tax Delinquency

The choropleth plot showing the spatial distribution of average property tax balances reveals distinct pocket of high tax burdens in Center City as well as South, North, Upper North, and Northeast Philadelphia.

```{r spatial-tax-map}
#| fig-dpi: 300
#| fig-height: 10
#| fig-width: 10

philly_tax_sf <- philly_tracts_sf %>%
  left_join(df_tax, by = "GEOID")

tax_map <- ggplot(philly_tax_sf) +
  geom_sf(aes(fill = avg_balance), color = NA) +
  scale_fill_viridis_c(
    option = "plasma",
    trans = "log1p",
    labels = dollar_format(),
    na.value = "gray50",
    breaks = c(1000, 5000, 10000, 50000)
  ) +
  labs(
    title = "Average Property Tax Delinquency by Census Tract",
    fill = "Average\nTax\nBalance",
    caption = "Data: OpenDataPhilly Real Estate Tax Balances"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.position = "right"
  )

tax_map
```

## 8. PRE-PANDEMIC BASELINE COMPARISON

### Monthly Filings vs Baseline Ratio

The plot and summary statistics shows that the Covid-19 pandemic posed a mass disruption to eviction trends as filings drop well below the pre-pandemic baseline before gradually increasing through 2021 and 2022.  By 2024 and 2025, filings stabilized near the pre-pandemic average, indicating a return to normative patterns. This trajectory emphasizes the need to capture immediate post-moratorium dynamics.

```{r monthly-baseline-ratio-plot}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 14
#| message: FALSE
#| warning: FALSE

monthly_baseline_ratio <- df_monthly %>%
  group_by(date) %>%
  summarize(
    total_filings = sum(filings_count, na.rm = TRUE),
    total_baseline = sum(filings_avg_prepandemic_baseline, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(total_baseline > 0) %>%
  mutate(ratio_to_baseline = total_filings / total_baseline)

# Ratio > 1 indicates system stress relative to historical norm.
baseline_ratio_plot <- ggplot(monthly_baseline_ratio, aes(x = date, y = ratio_to_baseline)) +
  geom_line(color = "#16A085", linewidth = 0.8, alpha = 0.6) +
  geom_smooth(method = "loess", se = TRUE, color = "#E74C3C", linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "black", linewidth = 1) +
  annotate("text",
    x = min(monthly_baseline_ratio$date + 100), y = 1.08,
    label = "Pre-Pandemic Average", hjust = 0, fontface = "bold"
  ) +
  geom_vline(xintercept = moratorium_start, linetype = "dotted", color = "#27AE60", linewidth = 1) +
  geom_vline(xintercept = moratorium_end, linetype = "dotted", color = "#E67E22", linewidth = 1) +
  labs(
    title = "Monthly Filings Normalized to Pre-Pandemic Baseline",
    x = "Date (Month)",
    y = "Monthly Filings / Pre-Pandemic Average Ratio",
    caption = "Data: Philadelphia Monthly Eviction Filings via Eviction Lab"
  ) +
  theme_eviction()

baseline_ratio_plot
```

```{r monthly-baseline-ratio-table}
monthly_baseline_ratio %>%
  mutate(period = case_when(
    date < moratorium_start  ~ "pre_moratorium",
    date <= moratorium_end   ~ "moratorium",
    date > moratorium_end    ~ "post_moratorium"
  )) %>%
  group_by(period) %>%
  summarize(
    n_weeks = n(),
    mean_ratio = mean(ratio_to_baseline, na.rm = TRUE),
    median_ratio = median(ratio_to_baseline, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  kable(
    digits = 2, caption = "Baseline Ratio Statistics by Policy Period",
    col.names = c("Period", "Weeks", "Mean Ratio", "Median Ratio")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# II. MODELING

## 1. FEATURE ENGINEERING FOR MODELING

### Creating Temporal Lag Variables

Temporal lag variables show how eviction filings carry over from one month to the next. If a tract had high filings last month, it is more likely to have high filings this month too. The one-, two-, and three-month lags capture short-term momentum, while the three-month rolling average smooths out short-term patterns.

```{r lagged-variables}
# When predicting month t, only data from months t-1, t-2, t-3, etc. is available.
df_model_prep <- df_monthly %>%
  filter(GEOID != "sealed", !is.na(GEOID), GEOID != "") %>%
  arrange(GEOID, date) %>%
  group_by(GEOID) %>%
  mutate(
    filings_lag1 = lag(filings_count, n = 1, order_by = date),
    filings_lag2 = lag(filings_count, n = 2, order_by = date),
    filings_lag3 = lag(filings_count, n = 3, order_by = date),
    # Rolling 3-month average of lagged values (t-1, t-2, t-3).
    # Original rollmean with align = "right" included current month, please don't change this.
    filings_ma3 = (filings_lag1 + filings_lag2 + filings_lag3) / 3
  ) %>%
  ungroup()

lag_coverage <- df_model_prep %>%
  summarize(
    total_obs = n(),
    missing_lag1 = sum(is.na(filings_lag1)),
    missing_lag2 = sum(is.na(filings_lag2)),
    missing_lag3 = sum(is.na(filings_lag3)),
    missing_ma3 = sum(is.na(filings_ma3)),
    pct_missing_lag1 = missing_lag1 / total_obs * 100,
    pct_missing_lag2 = missing_lag2 / total_obs * 100,
    pct_missing_lag3 = missing_lag3 / total_obs * 100,
    pct_missing_ma3 = missing_ma3 / total_obs * 100
  )

lag_coverage %>%
  kable(
    digits = 1, caption = "Lagged Variable Coverage Stats",
    col.names = c(
      "Observations", "Missing Lag 1", "Missing Lag 2", "Missing Lag 3",
      "Missing Lag 3-Month Rolling Average",
      "% Missing Lag 1", "% Missing Lag 2", "% Missing Lag 3",
      "% Missing Lag 3-Month Rolling Average"
    )
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Creating Inactivity Flag

About 37% of monthly tract observations have zero filings thus these flags are intended to help the model separate active periods from months with no eviction activity. This matters because it can potentially address problematic zero-inflation in the data and highlight longer stretches of inactivity that may reflect structural differences across tracts.

```{r add-zero-flags}
df_model_prep <- df_model_prep %>%
  group_by(GEOID) %>%
  mutate(
    was_zero_last_month = lag(if_else(filings_count == 0, 1, 0)),
    consecutive_zeros = sequence(rle(filings_count == 0)$lengths) * (filings_count == 0)
  ) %>%
  ungroup()

zero_coverage <- df_model_prep %>%
  summarize(
    total_obs = n(),
    missing_zero_last_month = sum(is.na(was_zero_last_month)),
    missing_consecutive_zeros = sum(is.na(consecutive_zeros)),
    pct_missing_zero_last_month = missing_zero_last_month / total_obs * 100,
    pct_missing_consecutive_zeros = missing_consecutive_zeros / total_obs * 100
  )

zero_coverage %>%
  kable(
    digits = 1, caption = "Zero Flag Variable Coverage Stats",
    col.names = c(
      "Observations", "Missing 0 Last Month", "Missing Consecutive Zeros",
      "% Missing 0 Last Month", "% Missing Consecutive Zeros"
    )
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Merging Claims Severity Data

The following represents various merges that integrate claims severity, tax delinquency, and American Community Survey (ACS) socioeconomic indicators into the tract-level eviction panel.

```{r merge-claims}
df_model_prep <- df_model_prep %>%
  mutate(year_month = floor_date(date, "month"))

df_claims_join <- df_claims %>%
  mutate(year_month = floor_date(date, "month")) %>%
  select(year_month, median_claim, claim_ratio)

df_model_prep <- df_model_prep %>%
  left_join(df_claims_join, by = "year_month")

claims_merge_stats <- df_model_prep %>%
  summarize(
    total_obs = n(),
    has_claims = sum(!is.na(median_claim)),
    pct_with_claims = has_claims / total_obs * 100
  )

cat(sprintf(
  "Observations with claims data: %s (%.1f%%)\n",
  comma(claims_merge_stats$has_claims),
  claims_merge_stats$pct_with_claims
))
```

### Merging Tax Delinquency Data

Tax delinquency serves as a proxy for landlord financial stress. Properties with outstanding tax balances may have landlords more likely to pursue evictions.

```{r merge-tax-data}
df_tax_join <- df_tax %>%
  select(
    GEOID, num_props, balance, avg_balance, log_delinquent_balance,
    delinquent_prop_count, delinquency_age_years
  )

df_model_prep <- df_model_prep %>%
  left_join(df_tax_join, by = "GEOID")

tax_merge_stats <- df_model_prep %>%
  summarize(
    total_obs = n(),
    has_tax_data = sum(!is.na(avg_balance)),
    pct_with_tax = has_tax_data / total_obs * 100
  )

cat(sprintf(
  "Observations with tax data: %s (%.1f%%)\n",
  comma(tax_merge_stats$has_tax_data),
  tax_merge_stats$pct_with_tax
))
```

### Integrating ACS Socioeconomic Data

American Community Survey data provides tract-level socioeconomic indicators that capture structural vulnerability to eviction. Variables chosen pose theoretical relevance to eviction data as they related to housing, income, and demographic composition. ACS data was further engineered to allow for meaningful interpretation and comparison with other variables.

```{r acs-data-retrieval}
# ACS 5-year estimates for Philadelphia County tracts (2023 vintage).
acs_vars <- c(
  # Housing characteristics.
  "B25003_001", # total occupied housing units
  "B25003_003", # renter-occupied units
  "B25064_001", # median gross rent
  "B25071_001", # median gross rent as % of income
  "B25077_001", # median home value

  # Income and poverty.
  "B19013_001", # median household income
  "B17001_001", # total population for poverty status
  "B17001_002", # population below poverty level

  # Employment.
  "B23025_001", # total population 16+ in labor force
  "B23025_005", # unemployed population

  # Education.
  "B15003_001", # total population 25+ for education
  "B15003_017", # high school diploma
  "B15003_022", # bachelor's degree
  "B15003_023", # master's degree
  "B15003_025", # doctorate degree

  # Demographics.
  "B01003_001", # total population
  "B01002_001", # median age

  # Household composition.
  "B11001_001", # total households
  "B11001_006", # female householder no spouse with children

  # Housing cost burden.
  "B25070_001", # total renter households for rent burden
  "B25070_007", # renters paying 30-34.9% income on rent
  "B25070_008", # renters paying 35-39.9% income on rent
  "B25070_009", # renters paying 40-49.9% income on rent
  "B25070_010"  # renters paying 50%+ income on rent
)

acs_data <- get_acs(
  geography = "tract",
  variables = acs_vars,
  state = "PA",
  county = "Philadelphia",
  year = 2023,
  survey = "acs5",
  output = "wide",
  geometry = FALSE
)

cat(sprintf("Tracts with ACS data: %d\n", nrow(acs_data)))
```

```{r acs-feature-engineering}
df_acs <- acs_data %>%
  transmute(
    GEOID = GEOID,
    pct_renter = if_else(B25003_001E > 0, B25003_003E / B25003_001E * 100, NA_real_),
    median_rent = B25064_001E,
    rent_burden_pct = B25071_001E,
    median_income = B19013_001E,
    log_median_income = log1p(B19013_001E),
    poverty_rate = if_else(B17001_001E > 0, B17001_002E / B17001_001E * 100, NA_real_),
    unemployment_rate = if_else(B23025_001E > 0, B23025_005E / B23025_001E * 100, NA_real_),
    pct_college = if_else(
      B15003_001E > 0,
      (B15003_022E + B15003_023E + B15003_025E) / B15003_001E * 100,
      NA_real_
    ),
    total_pop = B01003_001E,
    log_pop = log1p(B01003_001E),
    median_age = B01002_001E,
    pct_single_mother = if_else(B11001_001E > 0, B11001_006E / B11001_001E * 100, NA_real_),
    # Severe rent burden: renters paying 35%+ income on rent.
    severe_rent_burden = if_else(
      B25070_001E > 0,
      (B25070_008E + B25070_009E + B25070_010E) / B25070_001E * 100,
      NA_real_
    )
  )
```

Single mothers and those who experience severe rent burdens may be more theoretically linked to eviction filings as their means are limited.

```{r acs-summary-stats}
acs_summary <- df_acs %>%
  select(-GEOID) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarize(
    mean = mean(value, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE),
    min = min(value, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    pct_missing = sum(is.na(value)) / n() * 100,
    .groups = "drop"
  )

acs_summary %>%
  kable(
    digits = 2, caption = "ACS Feature Summary Stats",
    col.names = c("Variable", "Mean", "Median", "Standard Deviation", "Minimum", "Maximum", "% Missing")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r acs-merge}
df_model_prep <- df_model_prep %>%
  left_join(df_acs, by = "GEOID")

acs_merge_stats <- df_model_prep %>%
  summarize(
    total_obs = n(),
    has_acs = sum(!is.na(pct_renter)),
    pct_with_acs = has_acs / total_obs * 100
  )

cat(sprintf(
  "Observations with ACS data: %s (%.1f%%)\n",
  comma(acs_merge_stats$has_acs),
  acs_merge_stats$pct_with_acs
))
```

Creating a spatial weights matrix for our model is motivated by the maps predicting the spatial distribution of eviction filings across Philadelphia; eviction activity tends to not be isolated to specific tracts and emerge as part of broader neighborhood-level patterns. The spatial weights matrix allows the model to capture these localized spillovers, ensuring that predictions reflect the reality that housing instability is often similar in adjacent communities rather than occurring in isolation.

```{r spatial-weights}
tracts_in_data <- unique(df_model_prep$GEOID)

philly_model_sf <- philly_tracts_sf %>%
  filter(GEOID %in% tracts_in_data)

neighbors_queen <- poly2nb(philly_model_sf, queen = TRUE)
weights_matrix <- nb2listw(neighbors_queen, style = "W", zero.policy = TRUE)

cat(sprintf("Number of tracts in spatial analysis: %d\n", length(neighbors_queen)))
cat(sprintf(
  "Average number of neighbors per tract: %.1f\n",
  mean(card(neighbors_queen))
))
```

```{r spatial-lag-filings}
#| message: FALSE
#| warning: FALSE

# Use neighbors' filings from month t-1 (previous month) to avoid data leakage.
df_model_prep <- df_model_prep %>%
  arrange(GEOID, date) %>%
  group_by(GEOID) %>%
  mutate(filings_count_lag1 = lag(filings_count, n = 1)) %>%
  ungroup()

# For each month, calculate spatial lag using the lagged filings.
df_model_prep <- df_model_prep %>%
  group_by(date) %>%
  mutate(
    spatial_lag_filings = {
      sf_matched <- philly_model_sf %>%
        left_join(
          cur_data() %>% select(GEOID, filings_count_lag1),
          by = "GEOID"
        ) %>%
        mutate(filings_count_lag1 = replace_na(filings_count_lag1, 0))

      lag.listw(weights_matrix, sf_matched$filings_count_lag1, zero.policy = TRUE)
    }
  ) %>%
  ungroup()

cat(sprintf(
  "Observations with spatial lag: %s\n",
  comma(sum(!is.na(df_model_prep$spatial_lag_filings)))
))
```

### Creating Neighborhood Fixed-Effects

Neighborhoods provide a broader geographic context for understanding eviction filings. By linking tracts to neighborhood boundaries, we can test whether neighborhood characteristics are associated with filing rates. 

```{r neighborhood-boundaries}
neighborhoods <- st_read(path_neighbors) %>%
  st_make_valid() %>%
  st_transform(st_crs(philly_model_sf))

tracts_sf_subset <- philly_model_sf %>%
  select(GEOID, geometry)

neighborhoods_sf_subset <- neighborhoods %>%
  select(MAPNAME, geometry) %>%
  rename(neighborhood = MAPNAME)

# Keep tracts even if they don't perfectly intersect a neighborhood.
tracts_with_neighborhood <- st_join(
  tracts_sf_subset,
  neighborhoods_sf_subset,
  join = st_intersects,
  left = TRUE
) %>%
  st_drop_geometry()

# Take first match when a tract intersects multiple neighborhoods.
df_neighborhood_join <- tracts_with_neighborhood %>%
  select(GEOID, neighborhood) %>%
  group_by(GEOID) %>%
  slice(1) %>%
  ungroup()

df_model_prep <- df_model_prep %>%
  left_join(df_neighborhood_join, by = "GEOID")

cat(sprintf("Total observations: %s\n", comma(nrow(df_model_prep))))
cat(sprintf(
  "Observations with neighborhood data: %s\n",
  comma(sum(!is.na(df_model_prep$neighborhood)))
))

df_model_prep$neighborhood <- as.factor(df_model_prep$neighborhood)
df_model_prep$neighborhood <- relevel(df_model_prep$neighborhood, ref = "East Falls")
```

A binary flag was set to identify high concentrations of filings. This helps separate routine variation from extreme events and allows the model to account for rare but impactful spikes without letting them distort patterns in the broader dataset.

```{r cap-threshold}
cap_threshold <- 20

df_model_prep <- df_model_prep %>%
  mutate(is_extreme_spike = if_else(filings_count > cap_threshold, 1, 0))
```

We generated a correlation matrix for all numeric variables to identify redundancy and guide model specification. Because poverty rate and logged median income captured similar socioeconomic conditions, we retained poverty rate due to its stronger theoretical link to eviction risk. We selected delinquent property count over logged delinquent balance because it showed a stronger empirical correlation with eviction filings. Finally, we used the three-month rolling spatial lag  rather than the one- or two-month lags, as the rolling measure captured short-term fluctuations while reducing redundancy. This process ensured that the final predictors were theoretically relevant and minimally collinear.

```{r correlation-matrix}
#| fig-dpi: 300
#| fig-height: 20
#| fig-width: 20
#| message: FALSE
#| warning: FALSE

custom_colors <- colorRampPalette(c("#f03e37", "white", "#189ace"))
color_vector <- custom_colors(200)

df_model_prep_correlation <- df_model_prep[, sapply(df_model_prep, is.numeric)]

correlation_matrix <- cor(df_model_prep_correlation, use = "pairwise.complete.obs")

corrplot(correlation_matrix,
  method = "square",
  order = "FPC",
  type = "lower",
  diag = TRUE,
  tl.col = "black",
  cl.pos = "n",
  addCoef.col = "black",
  col = color_vector
)
```

## 2. TRAIN/TEST SPLIT STRATEGY

### Time-Based Validation Split

For a truly predictive model, we must use time-based splitting to avoid data leakage. The model learns from historical data and predicts future periods.

```{r train-test-split}
train_end_date <- as.Date("2023-12-31")
test_start_date <- as.Date("2024-01-01")

df_filtered <- df_model_prep %>%
  filter(
    !is.na(filings_ma3),
    !is.na(spatial_lag_filings),
    !is.na(delinquent_prop_count),
    !is.na(pct_renter),
    !is.na(poverty_rate),
    !is.na(severe_rent_burden),
    !is.na(pct_single_mother),
    !is.na(racial_majority),
    !is.na(moratorium_active),
    !is.na(neighborhood)
  )

# Cap target at ~99th percentile in training only for model stability.
df_train <- df_filtered %>%
  filter(date <= train_end_date) %>%
  mutate(filings_count_capped = pmin(filings_count, cap_threshold))

# Test set retains raw values for real-world validation.
df_test <- df_filtered %>%
  filter(date >= test_start_date)

cat("TRAIN/TEST SPLIT SUMMARY\n")
cat(sprintf("Training Period: %s to %s\n", min(df_train$date), max(df_train$date)))
cat(sprintf("Training Observations: %s\n", comma(nrow(df_train))))
cat(sprintf("Training Months: %d\n", n_distinct(df_train$date)))
cat(sprintf("Test Period: %s to %s\n", min(df_test$date), max(df_test$date)))
cat(sprintf("Test Observations: %s\n", comma(nrow(df_test))))
cat(sprintf("Test Months: %d\n", n_distinct(df_test$date)))
```

```{r outlier-cap}
# Cap lagged momentum signal at 99.8 percentile to reduce distortion from rare mass-eviction spikes.
df_train <- df_train %>% mutate(filings_ma3 = pmin(filings_ma3, 20))
df_test  <- df_test  %>% mutate(filings_ma3 = pmin(filings_ma3, 20))
```

```{r train-test-distribution}
train_test_comparison <- bind_rows(
  df_train %>% mutate(set = "training"),
  df_test %>% mutate(set = "test")
) %>%
  group_by(set) %>%
  summarize(
    n_obs = n(),
    mean_filings = mean(filings_count, na.rm = TRUE),
    median_filings = median(filings_count, na.rm = TRUE),
    sd_filings = sd(filings_count, na.rm = TRUE),
    zero_pct = sum(filings_count == 0) / n() * 100,
    .groups = "drop"
  )

train_test_comparison %>%
  kable(
    digits = 2, caption = "Filing Distribution Comparison: Training vs Test Sets",
    col.names = c("Set", "Observations", "Mean Filings", "Median Filings", "Standard Deviation Filings", "% Zero")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

To prevent mass displacement events from destabilizing the Negative Binomial model, we capped the target variable at the ~99th percentile with 20 filings in the training set only. However, we retained the raw, uncapped values in the test set to validate the model's performance against real-world unpredictability.

## 3. NEGATIVE BINOMIAL REGRESSION MODEL

### Model Specification Rationale

Because of overdispersion, Negative Binomial regression is preferred over Poisson.

### Model 1: Baseline with Temporal, Policy, and Monthly Features

```{r model-1}
model_nb_1 <- glm.nb(
  filings_count_capped ~ filings_ma3 + moratorium_active + factor(month_num) + is_extreme_spike,
  data = df_train
)

cat("BASELINE NEGATIVE BINOMIAL MODEL SUMMARY\n")
summary(model_nb_1)
```

### Model 2: Add Tax Delinquency and Spatial Lag

```{r model-2}
model_nb_2 <- glm.nb(
  filings_count_capped ~
    filings_ma3 + moratorium_active + factor(month_num) + is_extreme_spike +
    spatial_lag_filings + delinquent_prop_count,
  data = df_train
)

cat("2ND NEGATIVE BINOMIAL MODEL SUMMARY\n")
summary(model_nb_2)
```

### Model 3: Add ACS

```{r model-3}
model_nb_3 <- glm.nb(
  filings_count_capped ~
    filings_ma3 + moratorium_active + factor(month_num) + is_extreme_spike + neighborhood +
    spatial_lag_filings + delinquent_prop_count +
    pct_renter + severe_rent_burden + poverty_rate + pct_single_mother +
    factor(racial_majority),
  data = df_train
)

cat("3RD NEGATIVE BINOMIAL MODEL SUMMARY\n")
summary(model_nb_3)
```

### Model 4: Full Model with Interactions

```{r model-4}
model_nb_4 <- glm.nb(
  filings_count_capped ~
    filings_ma3 + moratorium_active + factor(month_num) + is_extreme_spike + neighborhood +
    spatial_lag_filings + delinquent_prop_count +
    pct_renter + severe_rent_burden + poverty_rate + pct_single_mother +
    factor(racial_majority) +
    factor(racial_majority) * moratorium_active,
  data = df_train
)

cat("FULL NEGATIVE BINOMIAL MODEL SUMMARY\n")
summary(model_nb_4)
```

```{r model-comparison}
model_comparison <- data.frame(
  model = c("Baseline", "+ Tax and Spatial", "+ ACS", "+ Interactions"),
  aic = c(AIC(model_nb_1), AIC(model_nb_2), AIC(model_nb_3), AIC(model_nb_4)),
  theta = c(model_nb_1$theta, model_nb_2$theta, model_nb_3$theta, model_nb_4$theta),
  loglik = c(logLik(model_nb_1), logLik(model_nb_2), logLik(model_nb_3), logLik(model_nb_4))
) %>%
  mutate(
    aic_change = aic - min(aic),
    rank = rank(aic)
  )

model_comparison %>%
  kable(
    digits = 2, caption = "Model Comparison Statistics (Lower AIC = Better Fit)",
    col.names = c("Model", "AIC", "Theta", "Log Likelihood", "AIC Change", "Rank")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The addition of the policy interaction term `(racial_majority * moratorium_active)` resulted in the lowest AIC, making it the best-fitting model. However, for deployment, the 3rd model is best with lower MAE and a more simple model to interpret. `filings_ma3` is a strong predicting for past filing volume and reinforcing that eviction risk is persistent. `is_extreme_spike` is significant showcasing tracts with a history of mass-eviction events face a structurally higher baseline risk at a whopping 5.8 IRR. `spatial_lag_filings` is positive and significant, showing that there is spatial spillover of eviction risk. `delinquent_prop_count` is also significant, and this reinforces the link between landlord financial instability and pressure to pursue eviction filings. The interaction terms also reveal a nuanced policy impact. While the moratorium reduced filings universally, the interaction coefficients indicate that the benefit was not uniform across racial groups, as what was expected.

## 4. MODEL VALIDATION AND DIAGNOSTICS

### Train and Test Predictions

Generate predictions for all four models on both training and test sets.

```{r predictions-all-models}
# Helper: add prediction columns to a data frame.
add_predictions <- function(df, model, target_col) {
  df %>%
    mutate(
      predicted = predict(model, newdata = ., type = "response"),
      residual = .data[[target_col]] - predicted,
      abs_error = abs(residual)
    )
}

# Training set predictions (against capped target).
train_pred_1 <- add_predictions(df_train, model_nb_1, "filings_count_capped")
train_pred_2 <- add_predictions(df_train, model_nb_2, "filings_count_capped")
train_pred_3 <- add_predictions(df_train, model_nb_3, "filings_count_capped")
train_pred_4 <- add_predictions(df_train, model_nb_4, "filings_count_capped")

# Test set predictions (against raw target).
test_pred_1 <- add_predictions(df_test, model_nb_1, "filings_count")
test_pred_2 <- add_predictions(df_test, model_nb_2, "filings_count")
test_pred_3 <- add_predictions(df_test, model_nb_3, "filings_count")
test_pred_4 <- add_predictions(df_test, model_nb_4, "filings_count")

cat(sprintf("Training Set: %s observations\n", comma(nrow(train_pred_1))))
cat(sprintf("Test Set: %s observations\n", comma(nrow(test_pred_1))))
```

### Training Set Performance

```{r training-metrics}
# Helper: compute standard metrics for a prediction data frame.
compute_metrics <- function(pred_df, label, target_col) {
  pred_df %>%
    summarize(
      model = label,
      n = n(),
      MAE = mean(abs_error, na.rm = TRUE),
      RMSE = sqrt(mean(residual^2, na.rm = TRUE)),
      mean_observed = mean(.data[[target_col]], na.rm = TRUE),
      mean_predicted = mean(predicted, na.rm = TRUE),
      correlation = cor(.data[[target_col]], predicted, use = "complete.obs"),
      bias = mean(predicted - .data[[target_col]], na.rm = TRUE)
    )
}

train_metrics <- bind_rows(
  compute_metrics(train_pred_1, "Baseline",       "filings_count_capped"),
  compute_metrics(train_pred_2, "+ Tax/Spatial",   "filings_count_capped"),
  compute_metrics(train_pred_3, "+ ACS",           "filings_count_capped"),
  compute_metrics(train_pred_4, "+ Interactions",  "filings_count_capped")
)

train_metrics %>%
  kable(
    digits = 3, caption = "Training Set Performance Across Models",
    col.names = c("Model", "Number", "MAE", "RMSE", "Mean Observed", "Mean Predicted", "Correlation", "Bias")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Test Set Performance

```{r test-metrics}
test_metrics <- bind_rows(
  compute_metrics(test_pred_1, "Baseline",       "filings_count"),
  compute_metrics(test_pred_2, "+ Tax/Spatial",   "filings_count"),
  compute_metrics(test_pred_3, "+ ACS",           "filings_count"),
  compute_metrics(test_pred_4, "+ Interactions",  "filings_count")
)

test_metrics %>%
  kable(
    digits = 3, caption = "Test Set Performance Across Models",
    col.names = c("Model", "Number", "MAE", "RMSE", "Mean Observed", "Mean Predicted", "Correlation", "Bias")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Overfitting Diagnostic Check

A critical concern in predictive modeling is overfitting, where the model learns noise in the training data rather than generalizable patterns. We assess this by comparing train vs test performance.

```{r overfitting-check}
overfitting_check <- train_metrics %>%
  select(model, MAE_train = MAE, RMSE_train = RMSE, corr_train = correlation) %>%
  left_join(
    test_metrics %>% select(model, MAE_test = MAE, RMSE_test = RMSE, corr_test = correlation),
    by = "model"
  ) %>%
  mutate(
    MAE_change = (MAE_test - MAE_train) / MAE_train * 100,
    RMSE_change = (RMSE_test - RMSE_train) / RMSE_train * 100,
    corr_change = (corr_train - corr_test) / corr_train * 100
  )

overfitting_check %>%
  kable(
    digits = 2, caption = "Train vs Test Performance: Overfitting Assessment",
    col.names = c(
      "Model", "Training MAE", "Training RMSE", "Training Correlation",
      "Testing MAE", "Testing RMSE", "Testing Correlation",
      "MAE Change", "RMSE Change", "Correlation Change"
    )
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The worst MAE degradation across models is 19.44%, so there is good generalization. In addition, the test MAE for the best model, the ACS addition, is off by 1.77 filings and accounts for 57% of variation. This is a good, foundational start considering this study is measuring a response variable that is very tied to human behavior.

### Select Best Model for Further Analysis

Based on the model comparison and overfitting assessment, we select the best-performing model for detailed diagnostics. Model 3 with ACS provides the best balance of fit and generalization and interpretation for officials.

```{r select-best-model}
# Model 3 selected for operational deployment.
df_test_pred  <- test_pred_3
df_train_pred <- train_pred_3

selected_test_metrics <- test_metrics %>% filter(model == "+ ACS")

cat("SELECTED MODEL: Model 3 (+ ACS)\n")
cat(sprintf("Test MAE: %.3f\n", selected_test_metrics$MAE))
cat(sprintf("Test RMSE: %.3f\n", selected_test_metrics$RMSE))
cat(sprintf("Test Correlation: %.3f\n", selected_test_metrics$correlation))
```

```{r predicted-vs-observed}
#| fig-dpi: 300
#| fig-height: 10
#| fig-width: 10
#| message: FALSE
#| warning: FALSE

pred_obs_plot <- df_test_pred %>%
  ggplot(aes(x = predicted, y = filings_count)) +
  geom_hex(bins = 50) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#E74C3C", linewidth = 1) +
  scale_fill_viridis_c(option = "plasma", trans = "log10", labels = comma) +
  scale_x_continuous(limits = c(0, 155)) +
  scale_y_continuous(limits = c(0, 50)) +
  labs(
    title = "Predicted vs Observed Eviction Filings (Test Set | Model 3)",
    subtitle = sprintf(
      "MAE = %.2f | RMSE = %.2f | Correlation = %.3f",
      selected_test_metrics$MAE,
      selected_test_metrics$RMSE,
      selected_test_metrics$correlation
    ),
    x = "Predicted Filings",
    y = "Observed Filings",
    fill = "Count",
    caption = "Data: Philadelphia Eviction Filings 2024-2025"
  ) +
  theme_eviction() +
  theme(legend.position = "right")

pred_obs_plot
```

### Performance by Racial Majority

```{r performance-by-race}
test_metrics_by_race <- df_test_pred %>%
  filter(!is.na(racial_majority), racial_majority != "") %>%
  group_by(racial_majority) %>%
  summarize(
    n = n(),
    mae = mean(abs_error, na.rm = TRUE),
    rmse = sqrt(mean(residual^2, na.rm = TRUE)),
    mean_observed = mean(filings_count, na.rm = TRUE),
    mean_predicted = mean(predicted, na.rm = TRUE),
    mean_error = mean(residual, na.rm = TRUE),
    .groups = "drop"
  )

test_metrics_by_race %>%
  kable(
    digits = 2, caption = "Test Set Performance by Tract Racial Majority",
    col.names = c("Racial Majority", "Number", "MAE", "RMSE", "Mean Observed", "Mean Predicted", "Mean Error")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The model, mirroring real-life eviction biases, also has a larger MAE for Black communities, with white communities having the lowest MAE.

### Residual Analysis

```{r residual-analysis}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 12
#| message: FALSE
#| warning: FALSE

residual_plot <- df_test_pred %>%
  ggplot(aes(x = residual)) +
  geom_histogram(bins = 50, fill = "#3498DB", color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#E74C3C", linewidth = 1) +
  scale_x_continuous(limits = c(-15, 15)) +
  labs(
    title = "Distribution of Prediction Residuals (Test Set)",
    subtitle = "Centered near zero indicates unbiased predictions.",
    x = "Residual (Observed - Predicted)",
    y = "Frequency",
    caption = "Data: Philadelphia Eviction Filings 2024-2025"
  ) +
  theme_eviction()

residual_plot
```

## 5. RISK CLASSIFICATION SYSTEM

### Creating Risk Quintiles

```{r risk-quintiles}
df_test_risk <- df_test_pred %>%
  mutate(
    risk_quintile = ntile(predicted, 5),
    risk_category = case_when(
      risk_quintile == 1 ~ "Lowest Risk",
      risk_quintile == 2 ~ "Low Risk",
      risk_quintile == 3 ~ "Moderate Risk",
      risk_quintile == 4 ~ "High Risk",
      risk_quintile == 5 ~ "Highest Risk"
    ),
    risk_category = factor(risk_category,
      levels = c("Lowest Risk", "Low Risk", "Moderate Risk", "High Risk", "Highest Risk")
    )
  )

risk_validation <- df_test_risk %>%
  group_by(risk_category) %>%
  summarize(
    n_tract_months = n(),
    mean_observed = mean(filings_count, na.rm = TRUE),
    median_observed = median(filings_count, na.rm = TRUE),
    total_filings = sum(filings_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(pct_total_filings = total_filings / sum(total_filings) * 100)

risk_validation %>%
  kable(
    digits = 2, caption = "Observed Filings by Predicted Risk Category",
    col.names = c("Risk Category", "Observations", "Mean Observed", "Median Observed", "Filings", "% Filings")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r risk-quintile-plot}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 10

risk_bar_plot <- risk_validation %>%
  ggplot(aes(x = risk_category, y = mean_observed, fill = risk_category)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", pct_total_filings)), vjust = -0.5, size = 4) +
  scale_fill_viridis_d(option = "plasma", direction = -1) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Mean Observed Filings by Predicted Risk Category",
    subtitle = "Higher predicted risk categories correctly capture higher observed filing rates.",
    x = "Predicted Risk Category",
    y = "Mean Observed Filings per Tract-Month",
    caption = "Data: Philadelphia Eviction Filings 2024-2025"
  ) +
  theme_eviction() +
  theme(legend.position = "none")

risk_bar_plot
```

### High-Risk Tract Profile

```{r high-risk-profile}
high_risk_profile <- df_test_risk %>%
  filter(risk_category == "Highest Risk") %>%
  group_by(racial_majority) %>%
  summarize(
    n_tract_months = n(),
    pct_of_high_risk = n() / nrow(df_test_risk %>% filter(risk_category == "Highest Risk")) * 100,
    mean_filings = mean(filings_count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_tract_months))

high_risk_profile %>%
  kable(
    digits = 1, caption = "Racial Composition of Highest Risk Tract-Months",
    col.names = c("Racial Majority", "Observations", "% High Risk", "Mean Filings")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## 6. MODEL INTERPRETATION

### Feature Importance

```{r feature-importance}
#| fig-dpi: 300
#| fig-height: 10
#| fig-width: 10

# Shared helper to categorize model terms.
categorize_term <- function(term) {
  case_when(
    str_detect(term, "filings_")                  ~ "Temporal",
    str_detect(term, "spatial_lag_filings")        ~ "Spatial",
    str_detect(term, "log_avg_balance")            ~ "Tax Delinquency",
    str_detect(term, "pct_renter|poverty_rate|severe_rent_burden|unemployment_rate|log_median_income|pct_single_mother") ~ "Socioeconomic (ACS)",
    str_detect(term, "factor\\(month_num\\)")      ~ "Seasonality",
    str_detect(term, "moratorium")                 ~ "Policy",
    str_detect(term, "factor\\(racial_majority\\)") ~ "Demographic",
    str_detect(term, "neighborhood")               ~ "Neighborhood",
    TRUE                                           ~ "Other"
  )
}

feature_importance <- tidy(model_nb_3) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    abs_estimate = abs(estimate),
    direction = if_else(estimate > 0, "Positive", "Negative"),
    category = categorize_term(term)
  ) %>%
  arrange(desc(abs_estimate)) %>%
  head(10)

importance_plot <- feature_importance %>%
  ggplot(aes(x = reorder(term, abs_estimate), y = estimate, fill = category)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Top 10 Most Important Predictors",
    subtitle = "Coefficient magnitude indicates effect size on log(expected filings).",
    x = "Predictor Variable",
    y = "Coefficient Estimate",
    fill = "Category",
    caption = "Negative Binomial Regression Coefficients"
  ) +
  theme_eviction() +
  theme(legend.position = "bottom")

importance_plot
```

```{r non-neighborhood-importance}
#| fig-dpi: 300
#| fig-height: 10
#| fig-width: 10

all_coefficients <- tidy(model_nb_3) %>%
  filter(
    term != "(Intercept)",
    !str_detect(term, "neighborhood")
  ) %>%
  mutate(
    abs_estimate = abs(estimate),
    direction = if_else(estimate > 0, "Positive", "Negative"),
    category = categorize_term(term)
  ) %>%
  arrange(desc(abs_estimate)) %>%
  head(10)

other_importance_plot <- all_coefficients %>%
  ggplot(aes(x = reorder(term, abs_estimate), y = estimate, fill = category)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "All Non-Neighborhood Predictors",
    subtitle = "Coefficient magnitude indicates effect size on log(expected filings).",
    x = "Predictor Variable",
    y = "Coefficient Estimate",
    fill = "Category"
  ) +
  theme_eviction() +
  theme(legend.position = "bottom")

other_importance_plot
```

```{r acs-feature-contribution}
acs_contribution <- tidy(model_nb_3) %>%
  filter(str_detect(term, "pct_renter|poverty_rate|severe_rent_burden|unemployment_rate")) %>%
  mutate(
    irr = exp(estimate),
    pct_change = (irr - 1) * 100,
    significant = if_else(p.value < 0.05, "Yes", "No")
  ) %>%
  rename(standard_error = std.error, p_value = p.value, percent_change = pct_change) %>%
  select(term, estimate, standard_error, p_value, irr, percent_change, significant)

acs_contribution %>%
  kable(
    digits = 3, caption = "ACS Socioeconomic Feature Effects",
    col.names = c("Term", "Estimate", "Standard Error", "P-Value", "IRR", "% Change", "Significant")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Incidence Rate Ratios

```{r incidence-rate-ratios}
irr_table <- tidy(model_nb_3) %>%
  mutate(
    irr = exp(estimate),
    irr_lower = exp(estimate - 1.96 * std.error),
    irr_upper = exp(estimate + 1.96 * std.error),
    pct_change = (irr - 1) * 100
  ) %>%
  filter(term != "(Intercept)") %>%
  rename(p_value = p.value) %>%
  select(term, irr, irr_lower, irr_upper, pct_change, p_value)

irr_table %>%
  head(15) %>%
  kable(
    digits = 3, caption = "Incidence Rate Ratios (IRR) for Key Predictors",
    col.names = c("Term", "IRR", "IRR Lower Confidence", "IRR Upper Confidence", "% Change", "P-Value")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

`filings_ma3`: For every one-unit increase in the 3-month moving average of past eviction filings, the expected number of current eviction filings increases by a factor of 1.114 or 11.4%. So a history of higher recent filings strongly predicts higher current filings.

`moratorium_active`: When an eviction moratorium is active the expected number of eviction filings is multiplied by 0.483 or is 51.7% lower compared to when the moratorium is not active.

`is_extreme_spike`: If a time period is classified as having an extreme spike in filings, the expected number of eviction filings is nearly six times higher multiplied by 5.808, or increases by a staggering 481%, compared to a period without a spike.

`neighborhoodAcademy Gardens`: Compared to the baseline East Falls neighborhood, Academy Gardens is expected to have 59.8% more eviction filings.

```{r operational-recommendation}
# Actionable output for city agencies with intervention thresholds.
operational_output <- df_test_pred %>%
  filter(predicted >= 5) %>%
  group_by(GEOID) %>%
  summarize(
    predicted_filings_next_month = mean(predicted),
    n_high_risk_months = n(),
    .groups = "drop"
  ) %>%
  left_join(df_acs, by = "GEOID") %>%
  arrange(desc(predicted_filings_next_month))

operational_output
```

By integrating temporal, spatial, and socioeconomic indicators, the best model achieves a prediction error of < 2 filings per tract, outperforming the baseline model. Also, the model successfully identifies the "Highest Risk" tracts that account for the vast majority of displacement. This tool offers the City of Philadelphia a concrete mechanism to shift from reactive to proactive, targeting aid to the specific blocks where it is needed most and ideally before any landlord-tenant tensions escalate into expensive legal battles.

# III. EQUITY ASSESSMENT

## 1. Impact Analysis

```{r equity-assessment}
#| fig-dpi: 300
#| fig-height: 7
#| fig-width: 12

equity_metrics <- df_test_pred %>%
  filter(!is.na(racial_majority), racial_majority != "") %>%
  group_by(racial_majority) %>%
  summarize(
    n_obs = n(),
    mean_observed = mean(filings_count, na.rm = TRUE),
    mean_predicted = mean(predicted, na.rm = TRUE),
    mae = mean(abs_error, na.rm = TRUE),
    mean_error = mean(residual, na.rm = TRUE),
    under_prediction_rate = sum(residual > 0) / n() * 100,
    .groups = "drop"
  )

equity_metrics %>%
  kable(
    digits = 2, caption = "Model Performance Metrics by Tract Racial Majority",
    col.names = c("Racial Majority", "Observations", "Mean Observed", "Mean Predicted", "MAE", "Mean Error", "Under-Prediction Rate")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r equity-plot-mae}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 12

equity_mae_plot <- equity_metrics %>%
  ggplot(aes(x = racial_majority, y = mae, fill = racial_majority)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.2f", mae)), vjust = -0.5, size = 4) +
  scale_fill_manual(values = racial_colors) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Model Accuracy (MAE) by Tract Racial Majority",
    subtitle = "Lower values indicate better prediction accuracy.",
    x = "Tract Racial Majority",
    y = "Mean Absolute Error",
    caption = "Data: Philadelphia Eviction Filings 2024-2025"
  ) +
  theme_eviction() +
  theme(legend.position = "none")

equity_mae_plot
```

## 2. Prediction Calibration by Race

```{r calibration-analysis}
#| fig-dpi: 300
#| fig-height: 8
#| fig-width: 12

calibration_plot <- df_test_pred %>%
  filter(!is.na(racial_majority), racial_majority != "") %>%
  mutate(predicted_bin = cut(predicted, breaks = seq(0, 15, by = 1), include.lowest = TRUE)) %>%
  filter(!is.na(predicted_bin)) %>%
  group_by(racial_majority, predicted_bin) %>%
  summarize(
    mean_predicted = mean(predicted, na.rm = TRUE),
    mean_observed = mean(filings_count, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>%
  filter(n >= 10) %>%
  ggplot(aes(x = mean_predicted, y = mean_observed, color = racial_majority)) +
  geom_point(aes(size = n), alpha = 0.75) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  scale_color_manual(values = racial_colors) +
  scale_size_continuous(range = c(2, 8)) +
  labs(
    title = "Model by Tract Racial Majority",
    subtitle = "Points on diagonal indicate good predictions.",
    x = "Mean Predicted Filings",
    y = "Mean Observed Filings",
    color = "Racial\nMajority",
    size = "Observations",
    caption = "Data: Philadelphia Eviction Filings 2024-2025"
  ) +
  theme_eviction() +
  theme(legend.position = "right")

calibration_plot
```

## 3. Risk Category Distribution by Race

```{r risk-by-race}
#| fig-dpi: 300
#| fig-height: 7
#| fig-width: 12

risk_by_race <- df_test_risk %>%
  filter(!is.na(racial_majority), racial_majority != "") %>%
  group_by(racial_majority, risk_category) %>%
  summarize(n = n(), .groups = "drop") %>%
  group_by(racial_majority) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

risk_race_plot <- risk_by_race %>%
  ggplot(aes(x = racial_majority, y = pct, fill = risk_category)) +
  geom_col(width = 0.7) +
  scale_fill_viridis_d(option = "plasma", direction = -1) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Risk Distribution by Tract Racial Majority",
    subtitle = "Black-majority tracts classified as high risk.",
    x = "Tract Racial Majority",
    y = "Percentage of Tract-Months",
    fill = "Risk\nCategory",
    caption = "Data: Philadelphia Eviction Filings 2024-2025"
  ) +
  theme_eviction() +
  theme(legend.position = "right")

risk_race_plot
```

```{r equity-plot}
#| fig-dpi: 300
#| fig-height: 6
#| fig-width: 12

equity_calibration_plot <- equity_metrics %>%
  pivot_longer(
    cols = c(mean_observed, mean_predicted),
    names_to = "type", values_to = "value"
  ) %>%
  mutate(type = if_else(type == "mean_observed", "Observed", "Predicted")) %>%
  ggplot(aes(x = racial_majority, y = value, fill = type)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("Observed" = "#E74C3C", "Predicted" = "#3498DB")) +
  labs(
    title = "Model Calibration by Racial Majority: Observed vs Predicted",
    subtitle = "Close alignment indicates equitable prediction accuracy across communities.",
    x = "Tract Racial Majority",
    y = "Mean Filings per Tract-Month",
    fill = "Type",
    caption = "Data: Philadelphia Eviction Filings 2024-2025"
  ) +
  theme_eviction() +
  theme(legend.position = "bottom")

equity_calibration_plot
```

Future modifications need to be made to account for Black communities who account for the majority of Philadelphia's population, despite predicting Black communities to be at more risk, there is the simultaneous higher MAE, meaning that this tool is fit as long as there is human oversight and it is not used for absolute decision-making. It struggles to capture the full issues that come from structural racism, so going into detail with the resource allocation it should be legal aid or cash assistance supportive survices, never property or police enforcement.

# SESSION INFORMATION

```{r session-info}
sessionInfo()
```

# DATA DOWNLOADS

[Eviction Lab Main Data](https://evictionlab.org/eviction-tracking/get-the-data/)

[Eviction Lab Claims Data](https://evictionlab.org/eviction-tracking/philadelphia-pa/)

[Real Estate Tax Balances](https://opendataphilly.org/datasets/real-estate-tax-balances/)

[Neighborhood Boundaries](https://opendataphilly.org/datasets/philadelphia-neighborhoods/)

[Tract Boundaries](https://opendataphilly.org/datasets/census-tracts/)

ACS 2023 Data via `tidycensus` API

# AI USAGE

**Free version of Claude Sonnet 4.5 used for debugging code, especially the final panel data portion.**

Several code chunks provided by Dr. Delmelle with adjustments from previous labs and lectures.
